---
title: "Parameterizing Model for Site-specific Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EIRprevmatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F, warning=F}
library(malariasimulation)
library(sf)
```

This vignette describes how to parameterize `malariasimulation` for site-specific seasonality and intervention parameters. 

<br>

## Seasonality

Seasonality is incorporated into `malariasimulation` through the `get_parameters()` function. 

```{r eval=FALSE}
params <- get_parameters(list(
  model_seasonality = TRUE,                 # model seasonal rainfall
  g0 = 0.284596,                            # rainfall fourier parameters
  g = c(-0.317878, -0.0017527, 0.116455),     
  h = c(-0.331361, 0.293128, -0.0617547))     
```

To calculate the g and h rainfall fourier parameters for a specific place and time, we can use the [umbrella package](https://github.com/mrc-ide/umbrella/tree/gee). To install `umbrella`, follow the instructions on the [package website](https://github.com/mrc-ide/umbrella/tree/gee) to install `rgee` and Google Earth Engine. 

```{r eval=FALSE}
# Add libraries
library(umbrella) # remotes::install_github('mrc-ide/umbrella@gee', force=T)
library(reticulate)
library(rgee) # remotes::install_github('r-spatial/rgee', force=T)

# Set python version (needs to be <=3.9!) and install earthengine --------------
py_discover_config() # check python version
reticulate::use_python('C:/Users/username/Anaconda3/python.exe', required=T)

# Alternatively use the Anaconda prompt to create a new environment
# conda create --name rgee python=3.9
# conda activate rgee
# conda install -c conda-forge earthengine-api

# install earthengine
rgee::ee_install_set_pyenv(py_path = "C:/Users/username/Anaconda3/envs/rgee/python.exe", 
                           py_env = 'rgee')

# check
ee_check() # may encounter an error about numpy not being installed, but keep going

# sign into google earth account
ee_Initialize(user='user_email', drive=T) 
```

<br>

Select an administrative boundary shapefile to use. These can be found and downloaded from [geoBoundaries](https://www.geoboundaries.org/index.html#getdata).

```{r, warning=F, message=F, fig.width=8, fig.height=4}
# import spatial data
admin1 <- readRDS("extdata/Kenya_admin1_simplified.rds")[,c(1,9)]

# select specific units for which you want rainfall data
sites <- admin1[which(admin1$shapeName %in% c('Bomet','Kericho')),]
  
# plot
plot(st_geometry(admin1), lwd = 2, col = 'khaki')
plot(st_geometry(sites), col = 'olivedrab4', add = T)
```

Set a time range and extract rainfall data from your sites using the `umbrella` package.

```{r, eval=F}
# specify time-range
start_date <- "2015-01-01"
end_date <- "2015-12-31"

# extract rainfall data
daily_rain <- pull_daily_rainfall(sf = sites, start_date = start_date, end_date = end_date)

# process
seasonality <- process(gee_output = daily_rain, shapeName) # rain data, grouping variable
```

Visualize rainfall patterns. The red dots represent daily rainfall in this location, and the blue line represents the fit to the data calculated using `umbrella`.

```{r, warning=F, message=F, fig.height=4, fig.width=8}
# read in seasonality data
seasonality <- readRDS("extdata/Kenya_rainfall.rds")

# take out data for Bomet
out <- do.call(data.frame, seasonality[1,]$profile)
out2 <- do.call(data.frame, seasonality[1,]$rainfall_data)
out2$t <- seq.int(nrow(out2))

# plot
plot(out2$t, out2$rainfall, type = "b", lty = 1, col = 'red',  
     xlab = 'days', ylab = 'rainfall (mm)', main = 'Daily rainfall in Bomet, Kenya, 2015')
lines(out$t * 365, out$y, type = "l", lty = 1, col = 'cornflowerblue')

```  
<br>

Resulting coefficients from each site can be printed from the model output. 

```{r}
# print coefficients
do.call(data.frame, seasonality[1,]$coefficients)
do.call(data.frame, seasonality[2,]$coefficients)
```

These coefficients can then be used in `malariasimulation'.
```{r, eval=F}
params <- get_parameters(list(
  human_population = 1000,
  model_seasonality = TRUE,   # assign seasonality
  g0 = 4.534334,                              
  g = c(-1.067149, -1.319041, 1.112425),     
  h = c(1.121452, -3.323766, -0.9617459),     
  prevalence_rendering_min_ages = 2 * 365,
  prevalence_rendering_max_ages = 10 * 365))

```

<br>

## Infection and clinical disease

[Malaria Atlas Project](https://malariaatlas.org/data-directory/)
 -  Plasmodium falciparum and P. vivax incidence count, incidence rate, parasite rate, and percentage change
 - P. falciparum mortality count and rate
 - Predicted P. falciparum parasite rate in 2-10 year olds in Africa, 2000-2015 (published in 2015)
 - Predicted P. falciparum incidence rate in Africa, 2000-2015 (published in 2015)
 - Predicted P. vivax relapse incidence per 100,000 person days globally, 2013 (published in 2012)
 - Predicted spatial limits of P. falciparum malaria transmission globally, 2010 (published 2011)
 - Predicted spatial limits of P. vivax malaria transmission globally, 2010 (published 2012)
 
[Malaria Indicator Surveys (MIS)](https://dhsprogram.com/)
 - Testing using microscopy and RDT among children 0-59 months
 - Country-wide, regional, or cluster-specific estimates
 
[MAP site-specific parasite rate database](https://malariaatlas.org/pr-survey-data/)
 - Data from cross-sectional household surveys, aggregated
 

## Interventions

[Demographic and Health Surveys (DHS)](https://dhsprogram.com/)
 - ITN ownership and use
 - Household exposure to IRS
 - Intermittent preventive treatment against malaria during pregnancy
 - Type and timing of treatment of high fever in children under 5

[Malaria Indicator Surveys (MIS)](https://dhsprogram.com/)
 - ITN ownership and use
 - Household exposure to IRS
 - Intermittent preventive treatment against malaria during pregnancy
 - Type and timing of treatment of high fever in children under 5
 - Diagnostic blood testing of children under 5 using microscopy and RDT

[Malaria Atlas Project](https://malariaatlas.org/data-directory/)
 - Predicted ACT coverage in Africa, 2000-2015 (published in 2015)
 - Predicted IRS coverage in Africa, 2000-2015 (published in 2015)
 - Predicted ITN coverage in Africa, 2000-2015 (published in 2015)

[Maps and metrics of insecticide-treated net access, use, and nets-per-capita in Africa from 2000-2020](https://www.nature.com/articles/s41467-021-23707-7)

[Indoor residual spraying for malaria control in sub-Saharan Africa 1997 to 2017: an adjusted retrospective analysis](https://malariajournal.biomedcentral.com/articles/10.1186/s12936-020-03216-6)

[Mapping trends in insecticide resistance phenotypes in African malaria vectors](https://journals.plos.org/plosbiology/article?id=10.1371/journal.pbio.3000633)

[Global estimation of anti-malarial drug effectiveness for the treatment of uncomplicated Plasmodium falciparum malaria 1991–2019](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7573874/)


## Vector composition

[Malaria Atlas Project](https://malariaatlas.org/data-directory/)
 - Predicted dominant malaria vector species globally, 2010
 - Predicted secondary dominant malaria vector species in Africa+, 2010
 
