---
title: "Modifying the carrying capacity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modifying the carrying capacity}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300,
  fig.width=7
)
```

```{r setup}
library(malariasimulation)
```

The vector model in malariasimulation initialises vector species
populations with a given carrying capacity. This capacity determines the
population size of mosquitoes that the environment can support,
ultimately feeding into the intensity of transmission in the simulation.

There are a number of reasons why the baseline carrying capacity may
change over time. Natural variations caused by seasonal rainfall lead to
intra-annual changes in the carrying capacity. Human intervention may
seek to reduce the carrying capacity, for example by implementing larval
source management. An invasive species may exploit a previously empty
niche, leading to an increase in the carrying capacity.

All of these phenomena can be captured in malariasimulation using a
selection of functions to modify the parameter list. These are
demonstrated below.

First we can set up the basic parameter list

```{r}
timesteps <- 365 * 2
p <- get_parameters(overrides = list(human_population = 5000)) |>
  set_equilibrium(init_EIR = 5)
```

## Seasonality

We can modify our basic parameter list to include a seasonal profile

```{r}
p_seasonal <- p
p_seasonal$model_seasonality = TRUE
p_seasonal$g0 = 0.28605
p_seasonal$g = c(0.20636, -0.0740318, -0.0009293)
p_seasonal$h = c(0.173743, -0.0730962, -0.116019)
p_seasonal <- p_seasonal |>
  set_equilibrium(init_EIR = 5)
```

and can run the simulations and compare the outputs

```{r, fig.width = 7, fig.height = 4, out.width='100%'}
set.seed(123)
s <- run_simulation(timesteps = timesteps, parameters = p)
s$pfpr <- s$n_detect_730_3650 / s$n_age_730_3650
set.seed(123)
s_seasonal <- run_simulation(timesteps = timesteps, parameters = p_seasonal)
s_seasonal$pfpr <- s_seasonal$n_detect_730_3650 / s_seasonal$n_age_730_3650

par(mfrow = c(1, 2))
plot(s$EIR_gamb ~ s$timestep, t = "l", ylim = c(0, 200), xlab = "Time", ylab = "EIR")
lines(s_seasonal$EIR_gamb ~ s_seasonal$timestep, col = "darkorchid3")
plot(s$pfpr ~ s$timestep, t = "l", ylim = c(0, 1), xlab = "Time", ylab = "PfPr")
lines(s_seasonal$pfpr ~ s_seasonal$timestep, col = "darkorchid3")
```

## Custom modification of the mosquito carrying capacity

In malariasimulation, we can also modify the carrying capacity over time in a more
bespoke manner. We can use the `set_carrying_capacity()` helper function to 
specify scalars that modify the species-specific carrying capacity through time,
relative to the baseline carrying capacity, at defined time steps. The baseline
carrying capacity is generated when calling `set_equilibrium()`, to match the 
specified EIR and population size.

This functionality allows us to model a number of useful things:

### Larval source management

The impact of larval source management (LSM) is captured by reducing the
carrying capacity. The helper function `set_carrying_capacity()`
can be used to specify this intervention

```{r, fig.width = 7, fig.height = 4, out.width='100%'}
# Specify the LSM coverage
lsm_coverage <- 0.8
# Set LSM by scaling the carrying capacity by (1 - coverage)
p_lsm <- p |>
  set_carrying_capacity(
    carrying_capacity_scalers = matrix(1 - lsm_coverage, ncol = 1),
    timesteps = 365
  )
set.seed(123)
s_lsm <- run_simulation(timesteps = timesteps, parameters = p_lsm)
s_lsm$pfpr <- s_lsm$n_detect_730_3650 / s_lsm$n_age_730_3650

par(mfrow = c(1, 2))
plot(s$EIR_gamb ~ s$timestep, t = "l", ylim = c(0, 150), xlab = "Time", ylab = "EIR")
lines(s_lsm$EIR_gamb ~ s_lsm$timestep, col = "darkorchid3")
abline(v = 365, lty = 2, col = "grey60")
plot(s$pfpr ~ s$timestep, t = "l", ylim = c(0, 1), xlab = "Time", ylab = "PfPr")
lines(s_lsm$pfpr ~ s_lsm$timestep, col = "darkorchid3")
abline(v = 365, lty = 2, col = "grey60")
```

### Invasive species

An invasive species may exploit a new niche, increase the carrying
capacity at the point of invasion. The function
`set_carrying_capacity()` gives complete freedom to allow
changes to the carrying capacity to be made. Here
we will demonstrate using carrying capacity rescaling to capture the
invasion of _Anopheles stephensi_.

```{r, fig.width = 7, fig.height = 4, out.width='100%'}
# First initialise the mosquito species so that the invasive species starts with
# a negligible, but non-zero proportion.
p_stephensi <- p |>
  set_species(list(gamb_params, steph_params), c(0.995, 1 - 0.995)) |>
  set_equilibrium(init_EIR = 5)

# Next, at the time point of invasion, we scale up the carrying capacity for
# the invasive species by a factor that will be dependent on the properties of
# the invasion.
p_stephensi <- p_stephensi |>
  set_carrying_capacity(
    carrying_capacity_scalers = matrix(c(1, 2000), ncol = 2),
    timesteps = 365
  )

set.seed(123)
s_stephensi <- run_simulation(timesteps = timesteps, parameters = p_stephensi)
s_stephensi$pfpr <- s_stephensi$n_detect_730_3650 / s_stephensi$n_age_730_3650

par(mfrow = c(1, 2))
plot(s_stephensi$EIR_gamb ~ s_stephensi$timestep,
     t = "l", ylim = c(0, 150), xlab = "Time", ylab = "EIR")
lines(s_stephensi$EIR_steph ~ s_stephensi$timestep, col = "darkorchid3")
abline(v = 365, lty = 2, col = "grey60")
plot(s_stephensi$pfpr ~ s_stephensi$timestep, t = "l", ylim = c(0, 1), xlab = "Time", ylab = "PfPr")
abline(v = 365, lty = 2, col = "grey60")
```

### Full flexibility 

The function `set_carrying_capacity()` can be used to modify or set very
specific carrying capacity changes over time. We can demonstrate this
below by arbitrarily stepping the carrying capacity up and down over
time

```{r, fig.width = 7, fig.height = 4, out.width='100%'}
p_flexible <- p |>
  set_carrying_capacity(
    carrying_capacity = matrix(c(0.1, 2, 0.1, 0.5, 0.9), ncol = 1),
    timesteps = seq(100, 500, 100)
  )

set.seed(123)
s_flexible <- run_simulation(timesteps = timesteps, parameters = p_flexible)
s_flexible$pfpr <- s_flexible$n_detect_730_3650 / s_flexible$n_age_730_3650

par(mfrow = c(1, 2))
plot(s$EIR_gamb ~ s$timestep, t = "l", ylim = c(0, 150), xlab = "Time", ylab = "EIR")
lines(s_flexible$EIR_gamb ~ s_flexible$timestep, col = "darkorchid3")
abline(v = 365, lty = 2, col = "grey60")
plot(s$pfpr ~ s$timestep, t = "l", ylim = c(0, 1), xlab = "Time", ylab = "PfPr")
lines(s_flexible$pfpr ~ s_flexible$timestep, col = "darkorchid3")
abline(v = 365, lty = 2, col = "grey60")
```
