---
title: "Matching PfPR2-10 to EIR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EIRprevmatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F, warning=F}
library(tidyverse)
library(malariasimulation)
library(mgcv)
library(fuzzyjoin)
```

Begin by parameterizing the model you wish to run. Be sure to include any seasonality factors and interventions you want to include at baseline. 

```{r parameters}
year <- 365
human_population <- 1000

params <- get_parameters(list(
  human_population = human_population,
  average_age = 8453.323,     # to match flat_demog
  model_seasonality = TRUE,   # assign seasonality
  g0 = 0.284596,                              
  g = c(-0.317878, -0.0017527, 0.116455),     
  h = c(-0.331361, 0.293128, -0.0617547),     
  prevalence_rendering_min_ages = 2 * year,    # set prev age range
  prevalence_rendering_max_ages = 10 * year,
  fvt = 0,
  v = 0,
  severe_enabled = T,
  individual_mosquitoes = FALSE))

  # set species / drugs / treatment parameters
params <- set_species(params, species = list(arab_params, fun_params, gamb_params), 
                      proportions = c(0.25, 0.25, 0.5))
params <- set_drugs(params, list(AL_params))
params <- set_clinical_treatment(params, 1, c(1), c(0.45))

```

Now run `malariasimulation` over a range of EIRs. The goal is to run enough points to generate a curve to which you can match PfPR2-10 to EIR effectively.

```{r model, message=F, warning=F}
# loop over malariasimulation runs
init_EIR <- c(0.01, 0.1, seq(1,9,1), seq(10, 50, by=5)) # set EIR values

# run model
outputs <- lapply(
  init_EIR,
  function(init) {
    p_i <- set_equilibrium(params, init)
    run_simulation(5 * year, p_i) # sim time = 5 years
  }
)

# output EIR values
EIR <- lapply(
  outputs,
  function(output) {
    mean(
      rowSums(
        output[
          output$timestep %in% seq(4 * 365, 5 * 365), # just use data from the last year
          grepl('EIR_', names(output))
        ] / human_population * year
      )
    )
  }
)

# output prev 2-10 values
prev <- lapply(
  outputs,
  function(output) {
    mean(
      output[
        output$timestep %in% seq(4 * 365, 5 * 365),
        'n_detect_730_3650'
      ] / output[
        output$timestep %in% seq(4 * 365, 5 * 365),
        'n_730_3650'
      ]
    )
  }
)

# create dataframe of initial EIR, output EIR, and prev 2-10 results
EIR_prev <- cbind(init_EIR, unlist(EIR), unlist(prev)) %>% 
  as_tibble() %>% rename(output_EIR = V2, prev = V3)

```

Now plot your results! Code is included to compare the results of matching PfPR2-10 to EIR based on `malariaEquilibrium` (blue line) versus matching based on parameterized `malariasimulation` runs (red line). Notice that the generated points do not form a smooth curve. We ran `malariasimulation` using a population of just 1,000. Increasing the population to 10,000 or even 100,000 will generate more accurate estimates, but will take longer to run. 

```{r plot, warning=F}
# calculate EIR / prev 2-10 relationship from malariEquilibrium
eir <- seq(from = 0.1, to = 50, by=.5)
eq_params <- malariaEquilibrium::load_parameter_set("Jamie_parameters.rds")

prev <- vapply( # calculate prevalence between 2:10 for a bunch of EIRs
  eir,
  function(eir) {
    eq <- malariaEquilibrium::human_equilibrium(
      eir,
      ft=0,
      p=eq_params,
      age=0:100
    )
    sum(eq$states[2:10, 'pos_M']) / sum(eq$states[2:10, 'prop'])
  },
  numeric(1)
)

prevmatch <- as_tibble(cbind(eir,prev))

colors <- c("malariaEquilibrium" = "blue", "malariasimulation" = "red")

p <- ggplot(data=EIR_prev) + 
  geom_point(aes(x=init_EIR, y=prev), color='black') + 
  stat_smooth(aes(x=init_EIR, y=prev, color='malariasimulation'), method = 'gam', n=1000) + 
  geom_line(data=prevmatch, aes(x=eir, y=prev, color = 'malariaEquilibrium')) + 
  labs(x='initial EIR', y=expression(paste(italic(Pf),"PR"[2-10])), color='model') + 
  scale_color_manual(values = colors) + 
  theme_classic()

p

```

Now extract values from the fitted line to generate EIR estimates at your desired PfPR2-10 values.

```{r table, warning=F}
# extract points from stat_smooth
p2 <- ggplot_build(p)
p2 <- p2$data[[2]]

p2 <- p2 %>% as_tibble() %>% select(x,y) %>% rename(init_EIR=x, pred=y) 

# Pre-intervention baseline PfPR2-10 starting at values 10, 25, 35, 55 
PfPR <- as_tibble_col(c(.10, .25, .35, .55), column_name="pfpr")


# match via stat_smooth predictions
match <- PfPR %>%
  fuzzyjoin::difference_left_join(p2, by=c("pfpr"="pred"), 
                                  max_dist=1, distance_col="dist") %>%
  group_by(pfpr) %>% slice_min(dist) %>% rename(PfPR = pfpr, EIR = init_EIR, predicted_PfPR = pred, difference = dist)

match
```
