---
title: "Forcing the EIR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forcing the EIR}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  dpi=300,
  fig.width=7
)
```

```{r setup}
library(malariasimulation)
```

By default, transmission in malariasimulation is an emergent property of a
fully-simulated mosquito population: the Entomological Inoculation Rate (EIR)
experienced by humans arises from the dynamics of the vector model. In some
situations it is instead useful to prescribe the EIR directly - for example when
fitting to an observed EIR series, or when exploring counterfactual transmission
scenarios that are specified rather than modelled.

The `set_forced_eir()` helper switches the model into this mode. When a forced
EIR is set, the mosquito component is not simulated. Instead, the number of
infectious bites delivered at each timestep is driven directly from the
user-supplied EIR, while the overdispersed distribution of bites among
individuals (heterogeneity by age and individual biting rate) is preserved.

The EIR is supplied as a step function, in units of infectious bites per **adult**
person per year - the same per-adult convention as the default `init_EIR`
argument of `set_equilibrium()`. Each value applies from its timestep until the
next change point (the EIR is 0 before the first change point, so set the first
timestep to `1` to force the EIR from the start). The total EIR is split across
mosquito species by their biting proportions.

Note the following:

* `set_equilibrium()` should still be used to initialise the human population at
  a sensible starting point. Because `set_forced_eir()` shares the per-adult EIR
  convention, forcing `EIR = X` reproduces a `set_equilibrium(init_EIR = X)` run.
* The prescribed EIR does not respond to anything that would ordinarily feed back
  through the vector population, so `set_forced_eir()` emits a warning. When
  combining a forced EIR with interventions: those acting **directly on humans**
  (drugs, MDA, vaccines) work as usual; the **personal-protection** effects of
  bednets and spraying still *redistribute* the fixed bites away from protected
  people (so do not double-count an intervention already baked into your
  prescribed EIR - supply the pre-intervention EIR); and interventions acting
  **only through the mosquito population**, such as `set_carrying_capacity()`,
  have no effect.
* There is no entomological lag, so responses to step changes appear a little
  earlier than in a mosquito-simulated run, and seasonality must be supplied
  directly as a fine-grained EIR series rather than via a rainfall model.

## A constant forced EIR

First we set up a basic parameter list, initialise the human population at
equilibrium, and then force the EIR to the same value.

```{r}
timesteps <- 365 * 3
init_EIR <- 20

p_forced <- get_parameters(overrides = list(human_population = 5000)) |>
  set_equilibrium(init_EIR = init_EIR) |>
  set_forced_eir(timesteps = 1, eir = init_EIR)
```

For comparison, we also run a standard simulation (with a simulated mosquito
population) at the same equilibrium EIR.

```{r}
p_standard <- get_parameters(overrides = list(human_population = 5000)) |>
  set_equilibrium(init_EIR = init_EIR)
```

Running both, the forced simulation tracks the standard simulation closely: by
initialising at equilibrium and forcing the EIR at the equilibrium value, the
prevalence remains near its equilibrium level.

```{r, fig.width = 7, fig.height = 4, out.width='100%'}
set.seed(123)
s_forced <- run_simulation(timesteps = timesteps, parameters = p_forced)
s_forced$pfpr <- s_forced$n_detect_lm_730_3650 / s_forced$n_age_730_3650
set.seed(123)
s_standard <- run_simulation(timesteps = timesteps, parameters = p_standard)
s_standard$pfpr <- s_standard$n_detect_lm_730_3650 / s_standard$n_age_730_3650

par(mfrow = c(1, 2))
plot(s_standard$pfpr ~ s_standard$timestep, t = "l", ylim = c(0, 1),
     xlab = "Time", ylab = "PfPr", main = "Prevalence")
lines(s_forced$pfpr ~ s_forced$timestep, col = "darkorchid3")
legend("topright", legend = c("standard", "forced"),
       col = c("black", "darkorchid3"), lty = 1, bty = "n")
```

Because the mosquito component is not simulated under a forced EIR, the
mosquito-state output columns (the `E`/`L`/`P`/`Sm`/`Pm`/`Im` counts, `total_M`,
`FOIM` and `mu`) are absent from the results. The EIR is still reported.

```{r}
# mosquito-state columns are present in the standard run ...
any(grepl("total_M", names(s_standard)))
# ... but absent in the forced run
any(grepl("total_M", names(s_forced)))
```

## A time-varying forced EIR

The main purpose of `set_forced_eir()` is to prescribe an EIR that changes over
time. Here we start at a low EIR and step it up sharply after the first year.

```{r, fig.width = 7, fig.height = 4, out.width='100%'}
p_tv <- get_parameters(overrides = list(human_population = 5000)) |>
  set_equilibrium(init_EIR = 1) |>
  set_forced_eir(timesteps = c(1, 365), eir = c(1, 100))

set.seed(123)
s_tv <- run_simulation(timesteps = 365 * 3, parameters = p_tv)
s_tv$pfpr <- s_tv$n_detect_lm_730_3650 / s_tv$n_age_730_3650

par(mfrow = c(1, 2))
plot(s_tv$EIR_gamb ~ s_tv$timestep, t = "l",
     xlab = "Time", ylab = "EIR", main = "Forced EIR")
plot(s_tv$pfpr ~ s_tv$timestep, t = "l", ylim = c(0, 1),
     xlab = "Time", ylab = "PfPr", main = "Prevalence")
```

Prevalence rises following the step increase in the forced EIR, illustrating how
an arbitrary, prescribed transmission history can be imposed on the human
population.
