---
title: "HRP2 Antigen Persistence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HRP2 Antigen Persistence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300,
  fig.width=7
)
```

```{r setup}
# Load the requisite packages:
library(malariasimulation)
# Set colour palette:
cols  <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

This vignette describes how `malariasimulation` models the persistence of the histidine-rich protein 2 (HRP2) antigen, and how to configure and interpret it.

HRP2-based rapid diagnostic tests (RDTs) are widely used to detect malaria, but HRP2 antigen can remain detectable in the blood for some time after a *P. falciparum* infection has actually cleared. This means HRP2-based prevalence estimates can overestimate the burden of current infection relative to methods that detect parasites directly, such as light microscopy (LM) or PCR. `malariasimulation` includes a simple model of this persistence so that simulated HRP2 prevalence can be compared directly with LM/PCR-based outputs, and with field data collected via RDTs.

*Note: HRP2 persistence is currently a placeholder feature - the default parameter values below are not yet calibrated against field data.*

## How HRP2 persistence is modelled

HRP2 positivity is tracked with a single binary state per person: are they currently HRP2 positive, or not? It responds to whether a person is *currently* infected:

* **Becoming positive.** Every new clinical infection (whether or not it goes on to be treated) is marked HRP2 positive immediately. New asymptomatic infections are marked positive with probability `hrp2_asymptomatic_prob * P(detectable by light microscopy)` - i.e. HRP2 is assumed to catch a subset of the asymptomatic infections that LM catches, not all of them (for *P. vivax*, asymptomatic infections are always assumed LM-detectable). Subpatent infections never produce detectable HRP2.
* **Staying positive.** While a person remains in any infected state (D, A, U or Tr) they stay HRP2 positive with no decay - they are still parasitaemic, so there is no reason for HRP2 to clear yet. Reinfection while already HRP2 positive has no additional effect.
* **Clearing.** Once a person's infection actually resolves (i.e. they reach state S), a constant per-timestep hazard of `1/hrp2_scale` is applied - the same exponential/memoryless convention used to clear infections elsewhere in the model (e.g. the `dd`/`da`/`du`/`dt` duration parameters). `hrp2_scale` is therefore the mean number of timesteps between an infection clearing and HRP2 becoming undetectable, and the same value is used for everyone, whether or not the resolved infection was treated.

Because clearance is keyed off the current disease state, this works identically regardless of whether recovery happened naturally, via clinical drug treatment, or via mass drug administration/chemoprevention (see [Mass Drug Administration and Chemoprevention](https://mrc-ide.github.io/malariasimulation/articles/MDA.html)).

The implied survival curve - the probability that someone is still HRP2 positive $x$ timesteps after their infection cleared - is simply:

$$P(\text{still positive}) = e^{-x/\texttt{hrp2\_scale}}$$

```{r, fig.align = 'center', out.width='100%'}
hrp2_scale <- 60 / log(2) # a 60-timestep half-life

x <- 0:300
survival <- exp(-x / hrp2_scale)

plot(x, survival, type = "l", col = cols[1], lwd = 2,
     xlab = "Timesteps since infection cleared (state S)",
     ylab = "P(still HRP2 positive)",
     ylim = c(0, 1), xaxs = "i", yaxs = "i")
grid(lty = 2, col = "grey80", lwd = 0.5)
```

## Parameters

HRP2 parameters are set using the `set_hrp2_parameters()` convenience function:

* `hrp2_scale`: the mean number of timesteps between an infection clearing (state S) and HRP2 becoming undetectable; default = 20 (placeholder, not yet calibrated).
* `hrp2_asymptomatic_prob`: the probability that a new asymptomatic infection which is detectable by light microscopy is also HRP2 positive (new clinical infections are always assumed HRP2 positive); default = 1 (placeholder, not yet calibrated).

```{r}
simparams <- get_parameters(list(human_population = 2000))
simparams <- set_hrp2_parameters(
  simparams,
  hrp2_scale = hrp2_scale,
  hrp2_asymptomatic_prob = 1
)
```

## Rendering HRP2 outputs

Two sets of outputs are rendered automatically whenever a simulation is run:

* `n_hrp2_positive`: the total number of individuals currently HRP2 positive, each timestep.
* `n_untreated_infections`/`n_treated_infections`: the number of new untreated and treated clinical infections that timestep (rendered as part of the same process as HRP2 updates, but this split is purely for reporting - it plays no role in HRP2 clearance itself).

Age-stratified versions of these outputs (`n_hrp2_positive_<lower>_<upper>`, `untreated_infections_<lower>_<upper>`, `treated_infections_<lower>_<upper>`) can be requested via `set_epi_outputs()`, in the same way as other age-stratified outputs such as LM/PCR prevalence (see `?set_epi_outputs`).

## Example: comparing HRP2 positivity to light microscopy prevalence

To illustrate the feature, we simulate a population with clinical treatment available, and compare HRP2 positivity to LM-detectable prevalence over time and by age group. We use `set_epi_outputs()` with matching age bands for both outputs so they share a denominator (`n_age_<lower>_<upper>`) and are directly comparable.

```{r}
year <- 365
sim_length <- 10 * year
human_population <- 2000
starting_EIR <- 10

simparams <- get_parameters(list(human_population = human_population))
simparams <- set_equilibrium(simparams, init_EIR = starting_EIR)
simparams <- set_drugs(simparams, list(AL_params))
simparams <- set_clinical_treatment(simparams, drug = 1, timesteps = 1, coverages = 0.5)
simparams <- set_hrp2_parameters(
  simparams,
  hrp2_scale = hrp2_scale,
  hrp2_asymptomatic_prob = 1
)
simparams <- set_epi_outputs(
  simparams,
  hrp2 = c(0, 5 * year, 100 * year),
  prevalence = c(0, 5 * year, 100 * year)
)

set.seed(1)
output <- run_simulation(timesteps = sim_length, parameters = simparams)
```

### Total HRP2 positivity over time

```{r, fig.align = 'center', out.width='100%'}
plot(output$timestep, output$n_hrp2_positive, type = "l", col = cols[1], lwd = 2,
     xlab = "Days", ylab = "Number HRP2 positive",
     xaxs = "i", yaxs = "i")
grid(lty = 2, col = "grey80", lwd = 0.5)
```

### HRP2 positivity by age group

```{r, fig.align = 'center', out.width='100%'}
plot(output$timestep, output$n_hrp2_positive_0_1824, type = "l", col = cols[1], lwd = 2,
     xlab = "Days", ylab = "Number HRP2 positive",
     ylim = range(c(output$n_hrp2_positive_0_1824, output$n_hrp2_positive_1825_36499)),
     xaxs = "i", yaxs = "i")
lines(output$timestep, output$n_hrp2_positive_1825_36499, col = cols[2], lwd = 2)
grid(lty = 2, col = "grey80", lwd = 0.5)
legend("topleft", legend = c("Under 5", "5+"), col = cols[1:2], lty = 1, lwd = 2, cex = 0.8)
```

### HRP2 vs. light microscopy prevalence

`n_hrp2_positive_<lower>_<upper>` and `n_detect_lm_<lower>_<upper>` share the same age-group denominator (`n_age_<lower>_<upper>`), so we can compare the two prevalence measures directly. Because D and Tr infections are always 100% LM-detectable and always HRP2 positive, the gap between the two measures is generated entirely by the asymptomatic (A) compartment, where LM applies an age- and immunity-dependent detection probability that HRP2 does not.

```{r, fig.align = 'center', out.width='100%'}
prev_hrp2_under5 <- output$n_hrp2_positive_0_1824 / output$n_age_0_1824
prev_lm_under5 <- output$n_detect_lm_0_1824 / output$n_age_0_1824
prev_hrp2_5plus <- output$n_hrp2_positive_1825_36499 / output$n_age_1825_36499
prev_lm_5plus <- output$n_detect_lm_1825_36499 / output$n_age_1825_36499

par(mfrow = c(1, 2))

plot(output$timestep, prev_hrp2_under5, type = "l", col = cols[1], lwd = 2,
     main = "Under 5", xlab = "Days", ylab = "Prevalence",
     ylim = range(c(prev_hrp2_under5, prev_lm_under5)), xaxs = "i", yaxs = "i")
lines(output$timestep, prev_lm_under5, col = cols[2], lwd = 2)
grid(lty = 2, col = "grey80", lwd = 0.5)
legend("topleft", legend = c("HRP2", "LM"), col = cols[1:2], lty = 1, lwd = 2, cex = 0.8)

plot(output$timestep, prev_hrp2_5plus, type = "l", col = cols[1], lwd = 2,
     main = "5+", xlab = "Days", ylab = "Prevalence",
     ylim = range(c(prev_hrp2_5plus, prev_lm_5plus)), xaxs = "i", yaxs = "i")
lines(output$timestep, prev_lm_5plus, col = cols[2], lwd = 2)
grid(lty = 2, col = "grey80", lwd = 0.5)
legend("topleft", legend = c("HRP2", "LM"), col = cols[1:2], lty = 1, lwd = 2, cex = 0.8)

par(mfrow = c(1, 1))
```

As expected, HRP2 positivity sits above LM-detectable prevalence in both age groups, and the gap is larger in the 5+ group: older/more-exposed individuals carry a larger asymptomatic reservoir, and acquired detection immunity makes LM increasingly likely to miss those asymptomatic infections - a discount that HRP2 positivity does not apply.

## Known limitation: initial conditions

`set_equilibrium()` seeds the initial disease-state distribution (S/A/D/U/Tr) to match steady-state prevalence for the configured EIR, but it does not seed HRP2 status to match - everyone starts HRP2 negative at `timestep = 0`, regardless of how many people `set_equilibrium()` has already placed into infected states. As a result, `n_hrp2_positive` starts at zero and only catches up to a realistic level over a short burn-in period (on the order of a few multiples of `hrp2_scale`), even though LM/PCR prevalence is already realistic from the first timestep. When comparing `n_hrp2_positive` to other prevalence measures, it is advisable to discard this initial burn-in period.
