---
title: "Model Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DiagrammeR)
```

This is an individual based model for P. Falciparum and malaria interventions.

# Human Biology

The human variables are documented in "R/variables.R".

The human biological processes are spread out between the following files:

 1. 'human_infection.R' (orange)
 2. 'disease_progression.R' (green)
 3. 'mortality_processes.R'(red)

```{r echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]        
  S
  D
  A
  U
  Tr
  
  node [shape = circle]
  i [label = '']
  
  # edge definitions with the node IDs
  S -> i [color='orange']
  i -> D [color='orange']
  i -> A [color='orange']
  i -> Tr [color='orange']
  D -> A -> U -> S [color='green']
  Tr -> s [color='green']
  A -> i [color='orange']
  U -> i [color='orange']
  U -> S [color='red']
  A -> S [color='red']
  D -> S [color='red']
  }")
```

# Mosquito Biology

The mosquito biological processes are spread out between the following files:

 1. 'src/mosquito_ode.cpp' (green)
 2. 'src/mosquito_emergence.cpp' (green)
 3. 'R/mosquito_biology.R' (orange)
 4. 'R/mosquito_biology.R' (red)

```{r echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  subgraph clusterode {
    style=filled;
    color=lightblue;
    node [shape = rectangle]
    E
    L
    P
    label='ODE'
  }

  node [shape = rectangle]
  Sm
  Pm
  Im
  NonExistent
  
  # edge definitions with the node IDs
  E -> L -> P [color='green', label='1']
  P -> Sm [color='green', label='2']
  Sm -> Pm -> Im [color='orange', label='3']
  Sm -> NonExistent [color='red', label='4']
  Pm -> NonExistent [color='red', label='4']
  Im -> NonExistent [color='red', label='4']
  }")
```
