---
title: "Model Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DiagrammeR)
```

This is an individual based model for P. Falciparum and malaria interventions.

#Human Biology

The human states and variables are documented in "R/individual.R".

The human biological processes are spread out between the following functions:

 1. 'create_infection_process' (orange)
 2. 'create_event_based_process' (green)
 3. 'create_mortality_process'(red)

```{r echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  node [shape = rectangle]        
  S
  D
  A
  U
  
  node [shape = circle]
  i [label = '']
  
  # edge definitions with the node IDs
  S -> i [color='orange']
  i -> D [color='orange']
  i -> A [color='orange']
  D -> A -> U -> S [color='green']
  A -> i [color='orange']
  U -> i [color='orange']
  U -> S [color='red']
  A -> S [color='red']
  D -> S [color='red']
  }")
```

#Mosquito Biology

The mosquito biological processes are spread out between the following functions:

 1. 'create_ode' (green)
 2. 'create_mosquito_emergence_process' (green)
 3. 'create_mosquito_infection_process' (orange)
 4. 'create_processes' (red)

```{r echo=FALSE}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = LR]
  
  subgraph clusterode {
    style=filled;
    color=lightblue;
    node [shape = rectangle]
    E
    L
    P
    label='ODE'
  }

  node [shape = rectangle]
  Sm
  Pm
  Im
  Unborn
  
  # edge definitions with the node IDs
  E -> L -> P [color='green', label='1']
  P -> Sm [color='green', label='2']
  Sm -> Pm -> Im [color='orange', label='3']
  Sm -> Unborn [color='red', label='4']
  Pm -> Unborn [color='red', label='4']
  Im -> Unborn [color='red', label='4']
  }")
```