---
title: "Non-malarial fevers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Non-malarial fevers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  fig.width = 7
)
```

```{r setup}
# Load the requisite package:
library(malariasimulation)
# Set colour palette:
cols <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Non‑malarial fevers (NMFs) can influence the number of routine malaria cases in two ways. They prompt individuals to seek treatment and, if an asymptomatic malaria infection is present, this can be detected and treated. NMFs therefore increase the number of infections that are diagnosed as part of routine care and drive antimalarial treatment.

In this vignette we demonstrate these effects in a highly seasonal setting. We assume a constant NMF rate across the year and show how this interacts with seasonal malaria transmission.

## Simulation setup

We will simulate three years in an area of high seasonal transmission. Treatment with artemether‑lumefantrine (AL) is available to 60% of clinical cases.

```{r}
# Time
year <- 365
years <- 3
sim_length <- years * year

# Set parameters for a high transmission seasonal area
p <- get_parameters(
  list(
    human_population = 10000,
    model_seasonality = TRUE,
    g0 = 0.284596,
    g = c(-0.317878, -0.0017527, 0.116455),
    h = c(-0.331361, 0.293128, -0.0617547)
  )
) |>
  set_drugs(list(AL_params)) |>
  set_clinical_treatment(
    drug = 1,
    timesteps = c(0),
    coverages = c(0.6)
  ) |>
  set_epi_outputs(
    clinical_incidence = c(0, 100 * year),
    severe_incidence = c(0, 100 * year)
  ) |>
  set_equilibrium(init_EIR = 50)

# Baseline run without NMFs
s1 <- run_simulation(sim_length, p)

# Run with NMFs (a single rate across all ages)
p2 <- set_nmf(p, ages = c(100 * year), rates = c(0.005))
s2 <- run_simulation(sim_length, p2)
```

## Aggregating incidence

NMF cases that are malaria‑positive are recorded in `n_nmf_malaria_detected`. We add these to the clinical incidence counts and aggregate by week.

```{r}
aggregate_week <- function(sim){
  sim$n_nmf_malaria_detected[is.na(sim$n_nmf_malaria_detected)] <- 0
  sim$n_inc_clinical_0_36499[is.na(sim$n_inc_clinical_0_36499)] <- 0
  sim$n_inc_clinical_0_36499 <- sim$n_inc_clinical_0_36499 + sim$n_nmf_malaria_detected
  week <- floor((sim$timestep - 1) / 7)
  agg <- aggregate(
    cbind(n_inc_clinical_0_36499, n_age_0_36499) ~ week,
    data = sim,
    sum
  )
  agg$clinical <- agg$n_inc_clinical_0_36499 / agg$n_age_0_36499 * 365
  agg$time <- agg$week * 7 / 365
  agg
}

r1 <- aggregate_week(s1)
r1$nmf <- "FALSE"

r2 <- aggregate_week(s2)
r2$nmf <- "TRUE"

r <- rbind(r1, r2)
```

## Visualisation

```{r, fig.align = 'center', out.width='100%'}
plot(
  r$time[r$nmf == "FALSE"], r$clinical[r$nmf == "FALSE"],
  type = "l", col = cols[1],
  xlab = "Time (years)", ylab = "Clinical incidence",
  ylim = range(r$clinical),
  xaxs = "i", yaxs = "i"
)
lines(
  r$time[r$nmf == "TRUE"], r$clinical[r$nmf == "TRUE"],
  col = cols[2]
)
legend(
  "topright", legend = c("Without NMF", "With NMF"),
  col = cols[1:2], lty = 1, bty = "n"
)
grid(lty = 2, col = "grey80", lwd = 0.5)
```

The impact of NMFs varies through the year because additional treatment coincides with the peaks and troughs of malaria transmission.

## Age‑specific NMF rates

Different age groups can experience different NMF rates. For example:

```{r}
p_age <- set_nmf(p, ages = c(5 * year, 15 * year), rates = c(0.01, 0.003))
```

This sets a higher rate for children under five and a lower rate for those aged five to fifteen, with no NMFs occurring after fifteen years of age.

## Limitations

The current implementation does not capture:

- seasonal variation in NMF rates,
- any correlation between malaria exposure and NMF risk, or
- age‑specific treatment seeking for NMFs.
```
