---
title: "Parameter Variation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter Variation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  
```{r setup, echo= FALSE}
# Load the requisite packages:
library(malariasimulation)
# Set colour palette:
cols <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300,
  fig.width=7
)
```

This vignette describes how variation in estimated model parameters can be incorporated into model simulations.

##  Estimating variation in parameters

The malariasimulation transmission model was fit utilizing a Bayesian framework, which produced a posterior distribution of parameter sets. Default parameters for the model were taken as the median across 50 random sets of parameter draws.

```{r, fig.align = 'center', out.width='100%'}
## plot parasite prevalence with default model parameters
simparams <- get_parameters(list(
  human_population = 100,
  individual_mosquitoes = FALSE
))

# Default (median) model parameters
sim <- run_simulation(timesteps = 1000, simparams)

# plot the default median parameter
plot(
  sim$timestep,
  sim$n_detect_730_3650 / sim$n_age_730_3650,
  t = "l",
  ylim = c(0, 1),
  ylab = "PfPr",
  xlab = "Time in days",
  xaxs = "i", yaxs = "i",
  lwd = 2, main = 'Parasite prevalence over time with default model parameters',
  cex.main = 0.9)
grid(lty = 2, col = "grey80", lwd = 0.5)
```

If needed, we can produce stochastic model outputs incorporating variation in model parameters. This is done with the `set_parameter_draw` function, which pulls parameter draws from this joint posterior of Markov chain Monte Carlo (MCMC) fitting. This function overrides the default model parameters with a sample from one of 1000 draws from the joint posterior.

Keep in mind that `set_parameter_draw` must be called prior to `set_equilibrium`, as the baseline transmission intensity must be calibrated to new model parameters.

```{r, fig.align = 'center', out.width='100%'}
## run simulation on different samples of the joint posterior distribution
# plot the default median parameter
plot(
  sim$timestep[1:500],
  sim$n_detect_730_3650[1:500] / sim$n_age_730_3650[1:500],
  t = "l",
  ylim = c(0, 1),
  ylab = "PfPr",
  xlab = "Time in days",
  xaxs = "i", yaxs = "i",
  main = 'Parasite prevalence over time for 8 sets of parameter draws',
  cex.main = 0.9
)
grid(lty = 2, col = "grey80", lwd = 0.5)

for (i in 1:7) {
  param_draw <- simparams |>
    set_parameter_draw(sample(1:1000, 1)) |>
    set_equilibrium(init_EIR = 5)
  
  sim <- run_simulation(timesteps = 500, param_draw)
  
  lines(sim$timestep, sim$n_detect_730_3650 / sim$n_age_730_3650, col = cols[i])
}
```


For more information on uncertainty in parameters, please refer to [The US President's Malaria Initiative, Plasmodium falciparum transmission and mortality: A modelling study](https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1002448), Supplemental material, Section 4. 