---
title: "P. vivax Model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P. vivax model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300,
  fig.width=7
)
```

```{r, output.lines=6}
# Load the requisite packages:
library(malariasimulation)
# Set colour palette:
cols  <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

This vignette describes the unique features of the *P. vivax* model.

## Using the *P. vivax* model

To select the *P. vivax* model, use the `parasite` argument in the `get_parameters()` function:

```{r, output.lines=6}
p <- get_parameters(parasite = "vivax")
```

Then we can run the simulation as normal:

```{r, output.lines=6}
simulation <- run_simulation(timesteps = 100, parameters = p)
```

## Parameters

Our default *P. vivax* parameters are sourced from a version of the analysis in White et al. 2018 (doi: 10.1038/s41467-018-05860-8), where model parameters were fitted to data from Papua New Guinea. The chosen parameter set fixes `b = 0.25` and `sigma_squared = 1.67` (for consistency with the *P. falciparum* model). The default parameters for both parasite species can be found in `data/raw/parasite_parameters.csv`, while parameters common to both models are given in `R/parameters.R` under `get_parameters()`.

Values for the model fitting posterior distribution can be selected using the `set_parameter_draw()` function as found in the [Parameter Variation](https://mrc-ide.github.io/malariasimulation/articles/ParameterVariation.html) vignette.

## Model structure

### Key Model References

White, M. T., Walker, P., Karl, S., Hetzel, M. W., Freeman, T., Waltmann, A., Laman, M., Robinson, L. J., Ghani, A., & Mueller, I. (2018). Mathematical modelling of the impact of expanding levels of malaria control interventions on Plasmodium vivax. Nature Communications, 9(1), 1–10. <https://doi.org/10.1038/s41467-018-05860-8>
