---
title: "Mosquito species"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mosquito species}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(malariasimulation)
```
    
### Plotting function
We will create a plotting function to visualise the output.
```{r}
cols <- c("#E69F00", "#56B4E9", "#009E73", 
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Plotting functions
plot_prev <- function() {
  plot(x = output_endophilic$timestep, y = output_endophilic$n_detect_730_3650 / output_endophilic$n_730_3650, 
       type = "l", col = cols[3], lwd = 2.5,
       xlab = "Time (days)", ylab = expression(paste(italic(Pf),"PR"[2-10])),
       xaxs = "i", yaxs = "i", bty = "l", ylim = c(0,1))
  lines(x = output_exophilic$timestep, y = output_exophilic$n_detect_730_3650 / output_exophilic$n_730_3650,
        col = cols[5], lwd = 2.5)
  abline(v = sprayingtimesteps, lty = 2, lwd = 2.5, col = "black")
  grid(lty = 2, col = "grey80", lwd = 1)
  legend("bottomleft", box.lty = 0, legend = c("Spraying", "Prevalence for endophilic mosquitoes", "Prevalence for exophilic mosquitoes"),
         col = c("black", cols[3], cols[5]), lty = c(2,1,1), lwd = 2.5)
}
```
    
As alluded to in the Model Structure, Vector Control: IRS, and Vector Control: Bed net vignettes, it is possible to account for varying proportions of mosquito species in the setting you are modelling using the `set_species()` function. IRS and bed nets could be expected to have different impacts depending on the proportion of each mosquito species because of variations in indoor resting, insecticide resistance, and proportion of bites on humans versus animals by species. If you have specified more than 1 species, then the arguments for `set_spraying()` and/or `set_bednets()` must be populated with values for each species at each timestep that the intervention is implemented. 

There are preset parameters for *An. gambiae*, *An. arabiensis*, and *An. funestus* that can be set by the helper functions `gamb_params()`, `arab_params()`, and `fun_params()`, respectively. The default values for each species in these helper functions are from [Sherrard-Smith et al., 2018](https://doi.org/10.1038/s41467-018-07357-w). The parameters are:  

* `blood_meal_rates`: the blood meal rates for each species 
* `foraging_time`: time spent taking blood meals    
* `Q0`: proportion of blood meals taken on humans  
* `phi_bednets`: proportion of bites taken in bed    
* `phi_indoors`: proportion of bites taken indoors    

We will demonstrate how to specify different mosquito species and how this could alter intervention impact using an example with IRS.

## IRS with single endophilic mosquito species 

Use the `get_parameters()` function to generate a list of parameters, accepting most of the default values, but modifying seasonality values to model a seasonal setting. We will use `set_species` to model mosquitoes similar to *An. funestus* but with a higher propensity to bite indoors. Then, we use the `set_equilibrium()` function to to initialise the model at a given entomological inoculation rate (EIR).  

We used the `set_spraying()` function to set an IRS intervention. This function takes as arguments the parameter list, timesteps of spraying, coverage of IRS in the population and a series of parameters related to the insecticide used in the IRS. The proportion of mosquitoes dying following entering a hut is dependent on the parameters `ls_theta`, the initial efficacy, and `ls_gamma`, how it changes over time. The proportion of mosquitoes successfully feeding is dependent on `ks_theta`, the initial impact of the insecticide in IRS, and `ks_gamma`, how the impact changes over time. Finally, the proportion of mosquitoes being deterred away from a sprayed hut depends on `ms_theta`, the initial impact of IRS, and `ms_gamma`, the change in impact over time. See a more comprehensive explanation in the Supplementary Information of [Sherrard-Smith et al., 2018](https://doi.org/10.1038/s41467-018-07357-w).

```{r}
year <- 365
month <- 30
sim_length <- 3 * year
human_population <- 1000
starting_EIR <- 50

simparams <- get_parameters(
  list(
    human_population = human_population,
    # seasonality parameters
    model_seasonality = TRUE, 
    g0 = 0.285277,
    g = c(-0.0248801, -0.0529426, -0.0168910),
    h = c(-0.0216681, -0.0242904, -0.0073646)
  )
)

peak <- peak_season_offset(simparams)

# Create an example mosquito species with a high value for `phi_indoors`
endophilic_mosquito_params <- fun_params
endophilic_mosquito_params$phi_indoors <- 0.9
endophilic_mosquito_params$species <- 'endophilic'

## Set mosquito species  with a high propensity for indoor biting
simparams <- set_species(
  simparams, 
  species = list(endophilic_mosquito_params),
  proportions = c(1)
)

sprayingtimesteps <- c(1, 2) * year + peak - 3 * month # There is a round of IRS in the 1st and second year 3 months prior to peak transmission.

simparams <- set_spraying(
  simparams,
  timesteps = sprayingtimesteps,
  coverages = rep(.8, 2), # # Each round covers 80% of the population 
  # nrows=length(timesteps), ncols=length(species) 
  ls_theta = matrix(2.025, nrow=length(sprayingtimesteps), ncol=1), # Matrix of mortality parameters per round of IRS and per species 
  ls_gamma = matrix(-0.009, nrow=length(sprayingtimesteps), ncol=1), # Matrix of mortality parameters per round of IRS and per species
  ks_theta = matrix(-2.222, nrow=length(sprayingtimesteps), ncol=1), # Matrix of feeding success parameters per round of IRS and per species
  ks_gamma = matrix(0.008, nrow=length(sprayingtimesteps), ncol=1), # Matrix of feeding success parameters per round of IRS and per species
  ms_theta = matrix(-1.232, nrow=length(sprayingtimesteps), ncol=1), # Matrix of deterrence parameters per round of IRS and per species
  ms_gamma = matrix(-0.009, nrow=length(sprayingtimesteps), ncol=1) # Matrix of deterrence parameters per round of IRS and per species
)

simparams <- set_equilibrium(simparams, starting_EIR)

# Running simulation with IRS
output_endophilic <- run_simulation(timesteps = sim_length, parameters = simparams)
```

We can see below that only the endophilic species is modelled. 
```{r}
simparams$species
simparams$species_proportions
```


## IRS with single exophilic mosquito species 

We will run the same model with IRS as above, but this time with an example mosquito species similar to *An. funestus*, but with a lower propensity to bite indoors. Note that now there are 3 columns for the `ls_theta`, `ls_gamma`, `ks_theta`, `ks_gamma`, `ms_theta`, and `ms_gamma` arguments. See the Vector Control: IRS vignette for more information about setting IRS with different mosquito species. 

```{r, fig.align="center", fig.height=4, fig.width=6}
# Create an example mosquito species with a low value for `phi_indoors`
exophilic_mosquito_params <- fun_params
exophilic_mosquito_params$phi_indoors <- 0.2
exophilic_mosquito_params$species <- 'exophilic'

## Set 3 species of mosquitoes, 10% An. arabiensis, 10% An. gambia, and 80% an example species with a lower propensity for indoor biting
simparams <- set_species(
  simparams, 
  species = list(exophilic_mosquito_params),
  proportions = c(1)
)

peak <- peak_season_offset(simparams)

sprayingtimesteps <- c(1, 2) * year + peak - 3 * month # There is a round of IRS in the 1st and second year 3 months prior to peak transmission.

simparams <- set_spraying(
  simparams,
  timesteps = sprayingtimesteps,
  coverages = rep(.8, 2), # # Each round covers 80% of the population 
  # nrows=length(timesteps), ncols=length(species) 
  ls_theta = matrix(2.025, nrow=length(sprayingtimesteps), ncol=1), # Matrix of mortality parameters
  ls_gamma = matrix(-0.009, nrow=length(sprayingtimesteps), ncol=1), # Matrix of mortality parameters per round of IRS and per species
  ks_theta = matrix(-2.222, nrow=length(sprayingtimesteps), ncol=1), # Matrix of feeding success parameters per round of IRS and per species
  ks_gamma = matrix(0.008, nrow=length(sprayingtimesteps), ncol=1), # Matrix of feeding success parameters per round of IRS and per species
  ms_theta = matrix(-1.232, nrow=length(sprayingtimesteps), ncol=1), # Matrix of deterrence parameters per round of IRS and per species
  ms_gamma = matrix(-0.009, nrow=length(sprayingtimesteps), ncol=1) # Matrix of deterrence parameters per round of IRS and per species
)

output_exophilic <- run_simulation(timesteps = sim_length, parameters = simparams)
```

### Plot adult female infectious mosquitoes by species over time
In the plot below, we can see that IRS is much more effective when the endophilic mosquito species is modelled compared to the scenario where an exophilic species is modelled. In this case, IRS will not be as effective because a larger proportion of bites take place outside of the home.  
```{r,fig.align="center", fig.height=5, fig.width=7}
plot_prev()
```

