---
title: "Treatment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Treatment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(malariasimulation)
```

# Introduction

This vignette describes how to implement clinical drug treatments in malariasimulation.

The malariasimulation package contains in-built parameters sets for three anti-malarial drugs:

1.  `AL_params`: artemether-lumefantrine (AL)
2.  `DHA_PQP_params`: dihydroartemisinin and piperaquine (DHA-PQP)
3.  `SP_AQ_params`: sulfadoxine-pyrimethamine and amodiaquine (SP-AQ)

While all these drugs can be used to treat malaria, DHA-PQP and SP-AQ remain in the body for some time following treatment, making them good candidates for chemoprevention in mass drug administrations (see [Mass Drug Administration](https://mrc-ide.github.io/malariasimulation/articles/MDA.html). Any of these drugs can be included in the parameter list using the `set_drugs()` function (see [drugs_parameters.R](https://github.com/mrc-ide/malariasimulation/blob/master/R/drug_parameters.R) for full parameter details).

Each drug parameter set is a vector of length four, that represent the drug efficacy, the infectiousness following treatment relative to an untreated infection, and parameters that determine the protection against reinfection ($P$): shape ($w$) and scale ($\lambda$). The decay of protection against infection follows a weibull distribution as follows:

$$P = e^{{(-t/\lambda)}^w}$$

The $P$ for each drug following treatment through time is plotted below.

```{r, echo = FALSE, fig.align = 'center', fig.height = 5, fig.width = 7}
cols  <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

calc_P <- function(w, lambda, t){
  P <- exp(-(t/lambda)^w)
}

t <- 1:70

P_matrix <- lapply(list(AL_params, DHA_PQP_params, SP_AQ_params), function(x){
  calc_P(w = x[3], lambda = x[4], t = t)
})

plot(x = t, y = P_matrix[[1]], type = "l", col = cols[1],
     xlab = "Days", ylab = expression(paste("Protection against reinfection (",italic(P),")")))
grid()

# 
invisible(sapply(2:3, function(x){
  points(x = t, y = P_matrix[[x]], type = "l", col = cols[x])
}))

# Add legend
legend("topright", legend = c("AL","DHA-PQP","SP-AQ"), col = cols, lty = 1)

```

For more details, please see:

Okell, L., Cairns, M., Griffin, J. *et al.* Contrasting benefits of different artemisinin combination therapies as first-line malaria treatments using model-based cost-effectiveness analysis. *Nat Commun* **5**, 5606 (2014). <https://doi.org/10.1038/ncomms6606>.

## Functions

Drug parameters can be incorporated into a complete parameter set using the `set_drugs` function which takes the full parameter set and a list of drug parameter sets.

A treatment regimen for each drug can then be described using the `set_clinical_treatment()` function which takes the drug index, a vector of timesteps at which a change in coverage occurs (where the initial coverage is 0 until the first timestep specified) and a vector of coverages for the drug that correspond with the timestep changes, as well as the complete parameter set.

Multiple drugs can be modelled simultaneously, with treatment coverage that can be specified for each drug. This function must be used for each drug included (e.g., if there are two drugs, `set_clinical_treatment()` must be called twice to specify the treatment plan).


# Example

We will run a simulation for one year using AL and DHA-PQP treatment regimens. We begin our AL treatment on day 100 at 40% coverage and provide this treatment for 200 days before the treatment stops. DHA-PQP treatment also has 40% coverage, but begins on day 200 and lasts for 100 days, also ending on day 300. Note that the sum of treatment coverages cannot exceed 100%.

Prior to the simulation, the function `set_equilibrium` must also be used to generate equilibrium values for the human and mosquito populations.

## Parameterisation and simulation

```{r}
# Daily simulation timesteps for two years
year <- 365
sim_length <- 1 * year

# With a population size of 1000
human_population <- 1000

# Set human population size
simparams <- get_parameters(overrides = list(human_population = human_population))

# Update parameter set with chosen drug-specific parameters (AL and DHA/PQP)
drug_params <- set_drugs(simparams, list(AL_params, DHA_PQP_params))

# Choose initial EIR to 10
starting_EIR <- 10 

# Set treatment program for AL (drug index = 1)
treatment_params <- set_clinical_treatment(
  parameters = drug_params,
  drug = 1,
  timesteps =  c(100,300), # Treatment coverage changes on day 100 and day 200
  coverages =  c(0.4,0)) # The initial treatment coverage (0%) is the default and does not need to be set

# Set treatment program for DHA-PQP (drug index = 2)
treatment_params <- set_clinical_treatment(
  parameters = treatment_params,
  drug =  2,
  timesteps = c(200,300),
  coverages =  c(0.3,0)) # Coverage as found in covs vector

# Use set_equilibrium to update the parameter set for a given initial EIR
treatment_params <- set_equilibrium(treatment_params, starting_EIR)

# Run simulation:
output <- run_simulation(sim_length, treatment_params)
  
```

## Visualisation

Following simulation of malaria transmission under this treatment regimen, we can now visualise the effect of the regimen on *Pf*PR~2-10~ through time using the `n_detect_730_3650` and `n_730_3650` outputs.

```{r, fig.align = 'center', fig.height = 5, fig.width = 7}
# Create a new variable for Pf Prevalence (Age 2-10)
output$PfPR2_10 <- output$n_detect_730_3650/output$n_730_3650

# Plot results
plot(x = output$timestep, y = output$PfPR2_10, type = "l",
     xlab = "Days", ylab = expression(paste(italic(Pf),"PR"[2-10])), col = cols[1])
  
# Add grid lines
grid()
```
