---
title: "Vaccines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
library(malariasimulation)
```

# Parameterisation

In this tutorial we are going to explore different strategies for malaria vaccine implementation. Begin by setting the default parameters to run the simulation from an equilibrium starting point.

```{r}
year <- 365
month <- 30
sim_length <- 3 * year
human_population <- 10000
starting_EIR <- 20

simparams <- get_parameters(list(
    human_population = human_population,
    incidence_rendering_min_ages = 0,
    incidence_rendering_max_ages = 5 * year,
    individual_mosquitoes = FALSE
  )
)

simparams <- set_equilibrium(simparams, starting_EIR)

```

Then we can run the simulation for a variety of vaccination strategies:

## Mass RTS,S

This intervention implements a round of RTS,S vaccination (3 doses) for individuals between 5 and 17 months at a single time point followed by a single booster dose 18 months after the primary series. We specify that the intervention should use the RTS,S vaccine through the `profile` and `booster_profile` inputs.

```{r}
rtssparams <- simparams

rtssparams <- set_mass_pev(
  rtssparams,
  profile = rtss_profile,
  timesteps = 1 * year,
  coverages = 1,
  min_wait = 0,
  min_ages = 5 * month,
  max_ages = 17 * month,
  booster_timestep = 18 * month,
  booster_coverage = 0.95,
  booster_profile = list(rtss_booster_profile)
)

output <- run_simulation(sim_length, rtssparams)
output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825

ggplot(data = output, aes(x = timestep / 365, y = clinical_incidence)) + 
  geom_line(col = "grey80") +
  stat_smooth(col = "darkblue", se = FALSE) +
  geom_vline(xintercept = 1, col = "red") +
  xlab("year") +
  ylab("Clinical incidence \n (per 1000 children aged 0-5)") +
  theme_bw()

```

We can look at the distribution of doses using the n_pev_mass_dose\_\* outputs:

```{r}
dose_data <- data.frame(timestep = output$timestep,
                        dose = rep(c("mass 1", "mass 2", "mass 3", "mass booster"), each = length(output$timestep)),
                        n = c(output$n_pev_mass_dose_1, output$n_pev_mass_dose_2, output$n_pev_mass_dose_3,
                              output$n_pev_mass_booster_1))
dose_data[dose_data$n > 0, ]

```

## RTS,S EPI

This intervention implements a more gradual age-based dosing of RTS,S vaccination such as would occur under the Expanded Programme on Immunization (EPI). Individuals here will be vaccinated once they reach 5 months of age. We specify that the intervention should use the RTS,S vaccine through the `profile` and `booster_profile` inputs.

```{r}
rtssepiparams <- simparams

# Add RTS,S strategy
rtssepiparams <- set_pev_epi(
  rtssepiparams,
  profile = rtss_profile,
  timesteps = 1 * year,
  coverages = 1,
  min_wait = 0,
  age = 5 * month,
  booster_timestep = 18 * month,
  booster_coverage = 0.95,
  booster_profile = list(rtss_booster_profile)
)

output <- run_simulation(sim_length * 2, rtssepiparams)
output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825

ggplot(data = output, aes(x = timestep / 365, y = clinical_incidence)) + 
  geom_line(col = "grey80") +
  stat_smooth(col = "darkblue", se = FALSE) +
  geom_vline(xintercept = 1, col = "red") +
  xlab("year") +
  ylab("Clinical incidence \n (per 1000 children aged 0-5)") +
  theme_bw()

```

We see a much more gradual decrease in clinical incidence following EPI implementation compared to the mass vaccination strategy.

## RTS,S seasonal boosters

In a seasonal setting, we can set booster timesteps relative to the start of the year. This allows us to consider seasonal dynamics and implement booster doses right before the start of the high transmission season to maximize impact. 

```{r}
rtssepiseasonalparams <- simparams
rtssepiseasonalparams$model_seasonality = TRUE
rtssepiseasonalparams$g0 = 0.28605
rtssepiseasonalparams$g = c(0.20636, -0.0740318, -0.0009293)
rtssepiseasonalparams$h = c(0.173743, -0.0730962, -0.116019)
peak <- peak_season_offset(rtssepiseasonalparams)

# Add RTS,S seasonal strategy
rtssepiseasonalparams <- set_pev_epi(
  rtssepiseasonalparams,
  profile = rtss_profile,
  timesteps = 1 * year,
  coverages = 1,
  min_wait = 6 * month,
  age = 5 * month,
  booster_timestep = (peak - 3 * month) + c(0, year),
  booster_coverage = rep(.7, 2),
  booster_profile = list(rtss_booster_profile, rtss_booster_profile),
  seasonal_boosters = TRUE
)

```

## RTS,S dosing

We can also try different dosing schedules using the `pev_doses` parameter. Here we administer dose 1 at 5 months, dose 2 30 days after dose 1, and dose 3 60 days after dose 1. We also administer three booster doses 12, 18, and 24 months following dose 3.  

```{r}
rtssepiparams2 <- rtssepiparams
rtssepiparams2$pev_doses <- c(0, 30, 60)
rtssepiparams2 <- set_pev_epi(
  rtssepiparams2,
  profile = rtss_profile,
  timesteps = 1 * year,
  coverages = 1,
  age = 5 * month,
  min_wait = 0,
  booster_timestep = c(12 * month, 18 * month, 24 * month),
  booster_coverage = c(1, 1, 1),
  booster_profile = list(
    rtss_booster_profile,
    rtss_booster_profile,
    rtss_booster_profile
  )
)
```

## TBV

We can also model vaccines with completely different modes of actions. For example, transmission blocking vaccines (TBV). This example demonstrates administration of 5 rounds of TBV to a proportion of the population ages 5 to 60 years.

```{r}
tbvparams <- simparams

tbvparams <- set_tbv(
  tbvparams,
  timesteps = round(c(1, 1.25, 1.5, 1.75, 2) * 365),
  coverages = rep(0.99, 5),
  ages = 5:60
)

output <- run_simulation(sim_length, tbvparams)
output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825

ggplot(data = output, aes(x = timestep / 365, y = clinical_incidence)) + 
  geom_line(col = "grey80") +
  stat_smooth(col = "darkblue", se = FALSE) +
  geom_vline(xintercept = 1, col = "red") +
  xlab("year") +
  ylab("Clinical incidence \n (per 1000 children aged 0-5)") +
  theme_bw()

```
