---
title: "Vaccines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
library(malariasimulation)
library(malariaEquilibrium)
library(reshape2)
```

# Parameterisation

We are going to set the default parameters to run the simulation from an equilibrium.

```{r}
year <- 365
sim_length <- 30 * year
human_population <- 1000
starting_EIR <- 50
repetitions <- 5

jamie_params <- load_parameter_set()
eq <- human_equilibrium(EIR = starting_EIR, ft = 0, p = jamie_params, age = 0:99)

simparams <- get_parameters(c(
  translate_jamie(remove_unused_jamie(jamie_params)),
  list(
    human_population = human_population,
    species = 'All',
    species_proportions = 1,
    model_seasonality = TRUE, # Let's try a bi-modal model
    g0 = 0.28605,
    g = c(0.20636, -0.0740318, -0.0009293),
    h = c(0.173743, -0.0730962, -0.116019),
    severe_enabled = TRUE,
    incidence_rendering_min_ages = 0,
    incidence_rendering_max_ages = 5 * year,
    severe_incidence_rendering_min_ages = 0,
    severe_incidence_rendering_max_ages = 5 * year
  )
))

simparams <- parameterise_human_equilibrium(simparams, eq)
simparams <- parameterise_mosquito_equilibrium(simparams, starting_EIR)

# Plotting functions
plot_prevalence <- function(output) {
  ggplot(output) + geom_line(
    aes(x = timestep, y = inc_0_1825, group = repetition, alpha = 1/repetitions)
  )
}

add_intervention_lines <- function(plot, events) {
  plot + geom_vline(
    data = events,
    mapping = aes(xintercept=timestep),
    color="blue"
  ) + geom_text(
    data = events,
    mapping = aes(x = timestep, y = 0, label = name),
    size = 4,
    angle = 90,
    vjust = -0.4,
    hjust = 0
  )
}
```

Then we can run the simulation for a variety of Vaccination strategies:

## RTS,S

This is a round of RTS,S vaccine for individuals between 5 - 17 months and a booster after 18 months; coverage of 80%; for 10 years:

```{r}
rtssparams <- simparams

peak <- peak_season_offset(rtssparams)
month <- 30

# Add RTS,S strategy
rtssevents = data.frame(
  timestep = c(10, 20) * year + peak - month, #vaccine efficacy kicks off a month before the peak
  name=c("RTS,S starts", "RTS,S ends")
)

rtssparams <- set_rtss(
  rtssparams,
  start = rtssevents$timestep[[1]],
  end = rtssevents$timestep[[2]],
  frequency = year,
  min_ages = 5 * month,
  max_ages = 17 * month,
  boosters = 18 * month,
  coverage = 0.8,
  booster_coverage = .7
)

output <- run_simulation_with_repetitions(
  sim_length,
  repetitions,
  rtssparams,
  parallel=TRUE
)
add_intervention_lines(plot_prevalence(output), rtssevents)
```

## TBV

This is a round of the TBV (transmission blocking) vaccine at ages 1, 2, 3 and 18; coverage of 80%; for 10 years:

```{r}
tbvparams <- simparams

# Add TBV strategy
tbvevents = data.frame(
  timestep = c(10, 20) * 365,
  name=c("TBV starts", "TBV ends")
)

tbvparams <- set_tbv(
  tbvparams,
  start = tbvevents$timestep[[1]],
  end = tbvevents$timestep[[2]],
  frequency = 365,
  ages = c(1, 2, 3, 18),
  coverage = 0.8
)

output <- run_simulation_with_repetitions(
  sim_length,
  repetitions,
  tbvparams,
  parallel=TRUE
)
add_intervention_lines(plot_prevalence(output), tbvevents)
```
