---
title: "Vaccines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

It is possible to model both pre-erythrocytic malaria vaccines and a theoretical transmission blocking vaccine (TBV). Parameters for the pre-erythrocytic vaccine can be set for a routine Expanded Programme on Immunization (EPI) strategy or a mass vaccination strategy using the helper functions `set_rtss_epi()` and `set_mass_rtss()`, while parameters for a TBV can be set using `set_tbv()`. Under an EPI strategy, everybody within a certain age range will be vaccinated within the routine immunisation system. A mass vaccination strategy is typically a one-time event where everybody, or everybody within a certain age group, is targeted for vaccination during a relatively short window of time. If you are modelling a seasonal setting, then it is also possible to schedule vaccination relative to the expected peak of malaria prevalence given by `peak_season_offset()`.

```{r setup, message=FALSE}
library(malariasimulation)

cols <- c("#E69F00", "#56B4E9", "#009E73", 
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

### Plotting functions

First, we will define a few functions to visualise the outputs.

```{r}
# Plotting clinical incidence 
plot_inci <- function(type = 'not seasonal'){
  if(type == 'seasonal'){
    comparison <- output_seas_control
    vaccinetime <- round(year + (peak - month * 3.5), 0) / 365
  } else {
    comparison <- output_control
    vaccinetime <- 1
  }
  output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825
  output$time_year <- output$timestep / year
  comparison$clinical_incidence <- 1000 * comparison$n_inc_0_1825 / comparison$n_0_1825
  comparison$time_year <- comparison$timestep / year

  plot(x = output$time_year, y = output$clinical_incidence, 
       type = 'l', col = cols[2],
       xlab = 'Time (years)', ylab = 'Clinical incidence (per 1000 children aged 0-5)',
       xaxs = "i", yaxs = "i", bty = "l", xaxs = "i", yaxs = "i")
  grid(lty = 2, col = 'grey80', lwd = 1)
  abline(v = vaccinetime, col = 'black', lty = 2, lwd = 2.5)
  curve_values <- loess(clinical_incidence ~ time_year, data = output, span = 0.3, method = 'loess') 
  lines(output$time_year, predict(curve_values),
        col = cols[5], lwd = 3)
  curve_valuescomp <- loess(clinical_incidence ~ time_year, data = comparison, span = 0.3, method = 'loess') 
  lines(comparison$time_year, predict(curve_valuescomp), 
        col = cols[6], lwd = 3)
  legend('topright', box.lty = 0, legend = c('Start of vaccination', 'Incidence for vaccine scenario', 'Smoothed incidence for \nno intervention scenario'),
         col = c('black', cols[5], cols[5]), lty = c(2,1,1), lwd = 2.5)
}

# Plot parasite prevalence
plot_prev <- function(type = 'not seasonal'){
    if(type == 'seasonal'){
    comparison <- output_seas_control
    vaccinetime <- round(year + (peak - month * 3.5), 0) / 365
  } else {
    comparison <- output_control
    vaccinetime <- 1
  }
  output$time_year <- output$timestep / year
  comparison$time_year <- comparison$timestep / year

  plot(x = output$time_year, y = output$n_detect_730_3650/output$n_730_3650, 
       type = 'l', col = cols[3], ylim=c(0,1),
       xlab = 'Time (years)', ylab = expression(paste(italic(Pf),"PR"[2-10])),
       xaxs = "i", yaxs = "i", bty = "l", xaxs = "i", yaxs = "i")
  grid(lty = 2, col = 'grey80', lwd = 1)
  lines(x = comparison$time_year, y = comparison$n_detect_730_3650/comparison$n_730_3650,
        col = cols[6])
  abline(v = vaccinetime, col = 'black', lty = 2, lwd = 2.5)
  legend('topright', box.lty = 0, legend = c('Start of vaccination', 'Prevalence for vaccine scenario', 'Prevalence for no intervention scenario'),
         col = c('black', cols[3], cols[6]), lty = c(2,1,1), lwd = 2.5)
}

# Plot dose timing
plot_doses <- function(){
  output$month <- ceiling(output$timestep / month)
  doses <- output[,c(2:5,38)]
  doses <- aggregate(cbind(doses[1:4]),
                      by = list(doses$month), 
                      FUN = sum)
  doses <- as.matrix(t(doses[,-1]))
  
  barplot(doses, xlab = "Month",
          ylab = 'Number of doses',
        col = cols[1:6],space = 0, 
        beside = FALSE, xaxs = "i", yaxs = "i")
  grid(lty = 2, col = 'grey80', lwd = 1)
  axis(side = 1, lty = 1, col = "black", pos = 0); axis(side = 2, lty = 1, col = "black")
  legend('topleft', box.lty = 0, legend = c('Dose 1','Dose 2','Dose 3','Booster 1'),
         col = cols[1:6], lty = rep(1, 4), lwd = 2.5, bg="transparent")
}
```

## Parameterisation

We will set the default parameters to run the simulation from an equilibrium that has no seasonality using the `set_equilibrium()` function.

```{r}
year <- 365
month <- 30
sim_length <- 3 * year
human_population <- 10000
starting_EIR <- 20

simparams <- get_parameters(list(
    human_population = human_population,
    incidence_rendering_min_ages = 0,
    incidence_rendering_max_ages = 5 * year,
    individual_mosquitoes = FALSE
  )
)

simparams <- set_equilibrium(simparams, starting_EIR)

# Run a model with no interventions in a setting with no seasonality
output_control <- run_simulation(sim_length * 2, simparams)
```

Then we can run the simulation for a variety of vaccination strategies.

## Mass RTS,S

First, we will model a single round of RTS,S vaccine given in a mass vaccination strategy where those within a given age range will be vaccinated to a specified coverage level and during a short window of time. In this example, individuals aged between 5 months and 10 years are vaccinated with a primary series, then with a booster 12 months after the initial vaccination. The age groups to be vaccinated as well as timing of all vaccinations can be modified within the `set_mass_rtss()` function.

```{r}
rtssparams <- set_mass_rtss(
  simparams,
  timesteps = 1 * year, # The single round of vaccination is at 1 year into the simulation.
  coverages = 1, # The vaccine is given to 100% of the population between the specified ages.
  min_wait = 0, # The minimum acceptable time since the last vaccination is 0.
  min_ages = 5 * month, # The minimum age for the target population to be vaccinated. 
  max_ages = 10 * year, # The maximum age for the target population to be vaccinated.
  boosters = 12 * month, # The booster is given at 12 months after the initial vaccination. 
  booster_coverage = 0.95 # Coverage of the booster dose is 95%.
)

output <- run_simulation(sim_length, rtssparams)
```

#### Plot clinical incidence and prevalence

```{r, fig.width = 8, fig.height = 4, fig.align = 'center'}
par(mfrow = c(1,2))
plot_inci()
plot_prev()
```

#### Plot doses

You can look at the distribution of doses using the `n_rtss_mass_dose_*` or `n_rtss_epi_dose_*` outputs. It is always a good idea to check the timing of vaccine doses to ensure that it is specified as intended.

```{r, fig.width = 6, fig.height = 4, fig.align = 'center', message=FALSE}
plot_doses()
```

### Seasonal mass vaccination

Mass vaccination can also be targeted seasonally. For example, we may want to have a mass vaccination campaign a few months prior to the peak transmission season. In the example below, we will create a parameter set with seasonality and first run a simulation with no vaccine, then run a simulation with a mass vaccination campaign to everyone between the ages of 5 months and 15 years.

```{r}
# Use the get_parameters() function to establish a parameter set with a seasonal profile with malaria incidence in children aged 0-5 rendered in the model output:
seas_simparams <- get_parameters(
	list(
    	human_population = human_population,
    	incidence_rendering_min_ages = 0,
    	incidence_rendering_max_ages = 5 * year,
    	individual_mosquitoes = FALSE,
    	model_seasonality = TRUE, # Let's try a bi-modal model
    	g0 = 0.285505,
    	g = c(-0.325352,-0.0109352,0.0779865),
    	h = c(-0.132815,0.104675,-0.013919)
  )
)

seas_simparams <- set_equilibrium(seas_simparams, starting_EIR)

# Run no vaccine scenario
output_seas_control <- run_simulation(sim_length *2, seas_simparams)
```

Next, we will run the scenario with a seasonal mass vaccination campaign.
```{r}
# Find the peak seasonality
peak <- peak_season_offset(seas_simparams)

seasmass_simparams <- set_mass_rtss(
  parameters = seas_simparams,
  timesteps = round(year + (peak - month * 3.5), 0),# The vaccination will begin 3.5 months prior to the peak seasonality in the second year.
  coverages = 1,
  min_ages = 5 * month,
  max_ages = 15 * year,
  min_wait = 0,
  boosters = round(c(12 * month + 2 * month)),
  booster_coverage = 1)

output <- run_simulation(sim_length * 2, seasmass_simparams)
```

#### Plot clinical incidence

```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_inci(type = 'seasonal')
```

#### Plot prevalence
```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_prev(type = 'seasonal')
```

#### Plot doses

```{r, fig.width = 6, fig.height = 4, fig.align = 'center', message=FALSE}
plot_doses()
```


## RTS,S EPI

You can also opt to vaccinate using the EPI strategy, where individuals will be vaccinated once they reach a certain age. In the example below, individuals will be vaccinated once they reach 5 months of age. For this intervention, we see a much more gradual impact following implementation.

```{r}
# Add RTS,S strategy
rtssepiparams <- set_rtss_epi(
  simparams,
  timesteps = 1 * year, # Vaccination will begin at 1 year.
  coverages = 1, # Vaccine coverage is 100%.
  min_wait = 0, # There is no minimum wait since the last vaccination.
  age = 5 * month, # Individuals will be vaccinated once they reach 5 months of age.
  boosters = 12 * month, # The booster is administered 12 months following the initial vaccination. 
  booster_coverage = 0.95 # 95% of those vaccinated with the primary series will be boosted.
)

output <- run_simulation(sim_length * 2, rtssepiparams)
```

#### Plot clinical incidence

```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_inci()
```

#### Plot prevalence
```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_prev()
```

#### Plot doses

```{r, fig.width = 6, fig.height = 4, fig.align = 'center', message=FALSE}
plot_doses()
```

### RTS,S seasonal boosters

In a seasonal setting, we can set booster timesteps relative to the start of the year instead of relative to the last dose. This allows us to target seasonal dynamics.

```{r}
# Calculate the peak of the transmission season based on seasonality parameters above.
peak <- peak_season_offset(seas_simparams) 

# Add RTS,S seasonal strategy
rtssepiseasonalparams <- set_rtss_epi(
  seas_simparams,
  timesteps = 1 * year, # Vaccination begins 1 year after the start of the simulation. 
  coverages = 1, # Vaccine coverage is 100%.
  min_wait = 6 * month, # When seasonal_boosters = TRUE, this is the minimum time between an individual receiving the final dose and the first booster.
  age = 5 * month, # Individuals will be vaccinated once they reach 5 months of age.
  boosters = peak - month * 3.5 , # Because seasonal_boosters = TRUE, the timestep here is relative to the start of the year. Here, we will give a booster at 3.5 months prior to peak transmission. 
  booster_coverage = c(.7), # 70% of the vaccinated population is boosted.
  seasonal_boosters = TRUE
)

output <- run_simulation(sim_length * 2, rtssepiseasonalparams)
```

#### Plot clinical incidence

```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_inci(type = 'seasonal')
```

#### Plot prevalence
```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_prev(type = 'seasonal')
```

#### Plot doses

```{r, fig.width = 6, fig.height = 4, fig.align = 'center', message=FALSE}
plot_doses()
```

### RTS,S dosing

You can try different dosing schedules using the `rtss_doses` parameter. As before, it is a good idea to visualise the timing of the vaccination strategy to ensure that you have implemented it as intended.

```{r}
rtssepiparams2 <- set_rtss_epi(
  simparams,
  timesteps = 1 * year,
  coverages = 1,
  age = 5 * month,
  min_wait = 0,
  boosters = c(12 * month, 24 * month), # Here, we are testing a strategy with 2 boosters.
  booster_coverage = c(1, 1)
)

rtssepiparams2$rtss_doses <- c(0, 30, 60) # setting the timesteps for the 3 doses in the primary series at 0, 1, 2 months

output <- run_simulation(sim_length * 2, rtssepiparams2)
```

#### Plot clinical incidence

```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_inci()
```

#### Plot prevalence
```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_prev()
```

#### Plot doses

```{r, fig.width = 6, fig.height = 4, fig.align = 'center', message=FALSE}
output$month <- ceiling(output$timestep / month)
doses <- output[,c(2:6,39)]
doses <- aggregate(cbind(doses[1:5]),
                    by = list(doses$month), 
                    FUN = sum)
doses <- as.matrix(t(doses[,-1]))

barplot(doses, xlab = "Month",
        ylab = 'Number of doses',
      col = cols[1:6],space = 0, 
      beside = FALSE, xaxs = "i", yaxs = "i")
grid(lty = 2, col = 'grey80', lwd = 1)
axis(side = 1, lty = 1, col = "black", pos = 0); axis(side = 2, lty = 1, col = "black")
legend('topleft', box.lty = 0, legend = c('Dose 1','Dose 2','Dose 3','Booster 1', 'Booster 2'),
       col = cols[1:6], lty = rep(1, 4), lwd = 2.5,bg="transparent")
```

## TBV

We can also model vaccines with completely different modes of actions. For example, a transmission blocking vaccine (TBV). This example shows 5 rounds of a TBV to everyone aged 5 years or older.

```{r}
tbvparams <- set_tbv(
  simparams, 
  timesteps = round(c(1, 1.25, 1.5, 1.75, 2) * 365), # The TBV will be given at year 1, year 2, and every 3 months in between.
  coverages = rep(0.99, 5), # 99% of the population will be covered.
  ages = 5:60 # The age range in years of those receiving the vaccine. 
)

output <- run_simulation(sim_length, tbvparams)
```

#### Plot clinical incidence

```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_inci()
```

#### Plot prevalence
```{r, fig.width = 6, fig.height = 4 fig.align = 'center'}
plot_prev()
```
