---
title: "Vaccines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vaccines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
library(malariasimulation)
```

# Parameterisation

We are going to set the default parameters to run the simulation from an equilibrium.

```{r}
year <- 365
month <- 30
sim_length <- 3 * year
human_population <- 10000
starting_EIR <- 20

simparams <- get_parameters(list(
    human_population = human_population,
    incidence_rendering_min_ages = 0,
    incidence_rendering_max_ages = 5 * year,
    individual_mosquitoes = FALSE
  )
)

simparams <- set_equilibrium(simparams, starting_EIR)

```

Then we can run the simulation for a variety of Vaccination strategies:

## Mass RTS,S

This is a round of RTS,S vaccine for individuals between 5 months and 10 years followed by a booster after 18 months:

```{r}
rtssparams <- simparams

rtssparams <- set_mass_rtss(
  rtssparams,
  timesteps = 1 * year,
  coverages = 1,
  min_wait = 0,
  min_ages = 5 * month,
  max_ages = 10 * year,
  boosters = 18 * month,
  booster_coverage = 0.95
)

output <- run_simulation(sim_length, rtssparams)
output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825

ggplot(data = output, aes(x = timestep / 365, y = clinical_incidence)) + 
  geom_line(col = "grey80") +
  stat_smooth(col = "darkblue", se = FALSE) +
  geom_vline(xintercept = 1, col = "red") +
  xlab("year") +
  ylab("Clinical incidence \n (per 1000 children aged 0-5)") +
  theme_bw()

```

You can look at the distribution of doses using the n_rtss_mass_dose\_\* outputs:

```{r}
dose_data <- data.frame(timestep = output$timestep,
                        dose = rep(c("mass 1", "mass 2", "mass 3", "mass booster"), each = length(output$timestep)),
                        n = c(output$n_rtss_mass_dose_1, output$n_rtss_mass_dose_2, output$n_rtss_mass_dose_3,
                              output$n_rtss_mass_booster_1))
dose_data[dose_data$n > 0, ]

```

## RTS,S EPI

You can opt for a more gradual dosing using the EPI strategy. 
Individuals will be vaccinated once they reach 5 months. For this intervention
we see a much more gradual increase in impact following implementation

```{r}
rtssepiparams <- simparams

# Add RTS,S strategy
rtssepiparams <- set_rtss_epi(
  rtssepiparams,
  timesteps = 1 * year,
  coverages = 1,
  min_wait = 0,
  age = 5 * month,
  boosters = 18 * month,
  booster_coverage = 0.95
)

output <- run_simulation(sim_length * 2, rtssepiparams)
output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825

ggplot(data = output, aes(x = timestep / 365, y = clinical_incidence)) + 
  geom_line(col = "grey80") +
  stat_smooth(col = "darkblue", se = FALSE) +
  geom_vline(xintercept = 1, col = "red") +
  xlab("year") +
  ylab("Clinical incidence \n (per 1000 children aged 0-5)") +
  theme_bw()
```

## RTS,S seasonal boosters

In a seasonal setting, we can set booster timesteps relative to the start of the year.
This allows us to target seasonal dynamics

```{r}
rtssepiseasonalparams <- simparams
rtssepiseasonalparams$model_seasonality = TRUE
rtssepiseasonalparams$g0 = 0.28605
rtssepiseasonalparams$g = c(0.20636, -0.0740318, -0.0009293)
rtssepiseasonalparams$h = c(0.173743, -0.0730962, -0.116019)
peak <- peak_season_offset(rtssepiseasonalparams)

# Add RTS,S seasonal strategy
rtssepiseasonalparams <- set_rtss_epi(
  rtssepiseasonalparams,
  timesteps = 1 * year,
  coverages = 1,
  min_wait = 6 * month,
  age = 5 * month,
  boosters = (peak - 3 * month) + c(0, year),
  booster_coverage = rep(.7, 2),
  seasonal_boosters = TRUE
)
```

## RTS,S dosing

You can try different dosing schedules using the rtss_doses parameter

```{r}
rtssepiparams2 <- rtssepiparams
rtssepiparams2$rtss_doses <- c(0, 30, 60)
rtssepiparams2 <- set_rtss_epi(
  rtssepiparams2,
  timesteps = 1 * year,
  coverages = 1,
  age = 5 * month,
  min_wait = 0,
  boosters = c(12 * month, 18 * month, 24 * month),
  booster_coverage = c(1, 1, 1)
)
```

## TBV

We can also model vaccines with completely different modes of actions. For example
a transmission blocking vaccines (TBV). This example shows 5 rounds of the TBV
to everyone aged over 5

```{r}
tbvparams <- simparams

tbvparams <- set_tbv(
  tbvparams,
  timesteps = round(c(1, 1.25, 1.5, 1.75, 2) * 365),
  coverages = rep(0.99, 5),
  ages = 5:60
)

output <- run_simulation(sim_length, tbvparams)
output$clinical_incidence <- 1000 * output$n_inc_0_1825 / output$n_0_1825

ggplot(data = output, aes(x = timestep / 365, y = clinical_incidence)) + 
  geom_line(col = "grey80") +
  stat_smooth(col = "darkblue", se = FALSE) +
  geom_vline(xintercept = 1, col = "red") +
  xlab("year") +
  ylab("Clinical incidence \n (per 1000 children aged 0-5)") +
  theme_bw()

```
