---
title: "Validation"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages(library(ggplot2))
library(malariasimulation)
library(malariaEquilibrium)
library(reshape2)
suppressPackageStartupMessages(library(cowplot))
```

# Equilibrium testing

This vignette demonstrates that the IBM is consistent with the [equilibrium solution](https://github.com/mrc-ide/malariaEquilibrium).

## Parameterising the model

Let's start by parameterising the model with the default parameters from the equilibirium solution code.

This involves:

 * translating parameters to their new names, and
 * removing parameters that are not used in the new model
 
```{r}
year <- 365
sim_length <- 50 * year
human_population <- 1000
repetitions <- 5

remove_keys <- function(x, n) { for (name in n) { x[[name]] <- NULL }; x }

jamie_params <- load_parameter_set("Jamie_parameters.rds")

params <- remove_keys(
  jamie_params,
  c(
    'rT', # to be included in treatment validation
    'rP', # to be included in treatment validation
    'tl', # unused!
    'aA', # used for microscopy and pcr calculations
    'aU', # used for microscopy and pcr calculations
    'tau',# unused!
    'f', # used later to calculate blood_meal_rates
    'Q0', # used later to calculate blood_meal_rates
    'cd_w', # unused!
    'cd_p', # unused!
    'cT' # to be included in treatment validation
  )
)

simparams <- translate_jamie(params)
simparams[['human_population']] <- human_population

#parameterise single species
simparams$variety_proportions <- 1
simparams$blood_meal_rates <- jamie_params$f * jamie_params$Q0

#use the ode vector model
simparams$vector_ode <- TRUE
```

Next we want to estimate the yearly EIR. This is an input parameter for the equilibrium solution, but for the IBM it's an emergent statistic...

```{r}

estimate_eir <- function() {
  output <- run_simulation(sim_length, simparams)
  sum(tail(output$mean_EIR, year))
}
```

We run our IBM model initialised with these state proportions to check for a divergence:

Let's first set up some utility functions:

```{r}
get_equilibrium <- function(EIR) {
  human_equilibrium(EIR = EIR, ft = 0, p = jamie_params, age = 0:99)
}

get_state_proportions <- function(eq) colSums(eq$states[,c('S', 'D', 'A', 'U')])

get_FOIM <- function(eq) eq$FOIM
```

And then run our models.

```{r}
estimated_eirs <- c()
outputs <- list()
densities <- c()
for (density in c(.5, 1, 5, 10, 50, 100, 500, 1000)) {
  simparams$density <- density
  EIR <- estimate_eir()
  if (EIR > 0) {
    densities <- c(densities, density)
    estimated_eirs <- c(estimated_eirs, EIR)
    eq <- get_equilibrium(EIR)
    simparams <- parameterise_equilibrium(get_state_proportions(eq), simparams)
    output <- run_simulation_with_repetitions(
      sim_length,
      repetitions,
      simparams,
      parallel=TRUE
    )
    outputs[[length(outputs) + 1]] <- output
  }
}
```

Plot EIR vs Prev

```{r}
eirs <- vapply(
  outputs,
  function(output) {
    mean(vapply(
      seq(repetitions),
      function(r) {
        sum(tail(subset(output, output$repetition == r)$mean_EIR, year))
      },
      numeric(1)
    ))
  },
  numeric(1)
)

prevs <- vapply(
  outputs,
  function(output) {
    mean(vapply(
      seq(repetitions),
      function(r) {
        mean(
          tail(
            subset(output, output$repetition == r)$prev_2_10,
            year
          ) / tail(
            subset(output, output$repetition == r)$n_2_10,
            year
          )
        )
      },
      numeric(1)
    ))
  },
  numeric(1)
)

ggplot(
  data.frame(eir = eirs, prev = prevs)
) + geom_line(
  aes(x = eir, y = prev)
) + scale_x_continuous(trans='log10')
```

Plot agreement

```{r, fig.height = 12}

agreement_plots <- lapply(
  seq_along(outputs),
  function(i) {
    output <- outputs[[i]]
    pop_size <- simparams$human_population
    state_props <- get_state_proportions(get_equilibrium(estimated_eirs[[i]]))
    equilibrium <- data.frame(
      timestep = seq(sim_length),
      human_S_count = rep(state_props[['S']] * pop_size, sim_length),
      human_D_count = rep(state_props[['D']] * pop_size, sim_length),
      human_A_count = rep(state_props[['A']] * pop_size, sim_length),
      human_U_count = rep(state_props[['U']] * pop_size, sim_length)
    )
    ggplot(
      melt(
        output[c(
          'timestep',
          'repetition',
          'human_S_count',
          'human_D_count',
          'human_A_count',
          'human_U_count'
        )],
        id.vars=c('timestep', 'repetition')
      )
    ) + geom_line(
      aes(
        x = timestep,
        y = value,
        group = interaction(variable, repetition),
        color = variable,
        alpha = 1/repetitions
      )
    ) + geom_line(
      data = melt(equilibrium, 'timestep'),
      aes(x = timestep, y = value, group = variable, color = variable),
      linetype = 'dashed'
    )
  }
)

# consolidate legends
legend <- get_legend(agreement_plots[[1]])
agreement_plots <- lapply(
  agreement_plots,
  function(plot) plot + theme(legend.position="none")
)
plot_grid(
  plot_grid(
    plotlist = agreement_plots,
    labels = densities,
    ncol = 2
  ),
  legend,
  rel_widths = c(1, .1)
)
```

Immunity

```{r, fig.height = 12}
immunity_plots <- lapply(
  outputs,
  function(output) {
    ggplot(
      melt(
        output[c(
          'timestep',
          'repetition',
          'human_ICA_mean',
          'human_ICM_mean',
          'human_IB_mean',
          'human_ID_mean'
        )],
        c('timestep', 'repetition')
      )
    ) + geom_line(
      aes(
        x = timestep,
        y = value,
        group = interaction(variable, repetition),
        color = variable,
        alpha = 1/repetitions
      )
    )
  }
)

# consolidate legends
legend <- get_legend(immunity_plots[[1]])
immunity_plots <- lapply(
  immunity_plots,
  function(plot) plot + theme(legend.position="none")
)
plot_grid(
  plot_grid(
    plotlist = immunity_plots,
    labels = densities,
    ncol = 2
  ),
  legend,
  rel_widths = c(1, .1)
)
```

EIR

```{r, fig.height = 12}
eir_plots <- lapply(
  seq_along(outputs),
  function(i) {
    output <- outputs[[i]]
    EIR <- estimated_eirs[[i]]
    ggplot(
      output[c(
          'timestep',
          'repetition',
          'mean_EIR'
        )]
    ) + geom_line(
      aes(x = timestep, y = mean_EIR, group = repetition, alpha = 1/repetitions)
    ) + geom_line(
        data = data.frame(timestep=seq(sim_length), EIR=rep(EIR/365, sim_length)),
        aes(x = timestep, y = EIR),
        linetype = 'dashed'
    ) + theme(legend.position="none")
  }
)

plot_grid(
  plotlist = eir_plots,
  labels = densities,
  ncol = 2
)
```

FOIM

```{r, fig.height = 12}
foim_plots <- lapply(
  seq_along(outputs),
  function(i) {
    output <- outputs[[i]]
    FOIM <- get_FOIM(get_equilibrium(estimated_eirs[[i]]))
    ggplot(
      output[c(
          'timestep',
          'repetition',
          'FOIM_1'
        )]
    ) + geom_line(
      aes(x = timestep, y = FOIM_1, group = 'repetition', alpha=1/repetitions)
    ) + geom_line(
        data = data.frame(timestep=seq(sim_length), FOIM_1=rep(FOIM, sim_length)),
        aes(x = timestep, y = FOIM_1),
        linetype = 'dashed'
    ) + theme(legend.position="none")
  }
)

plot_grid(
  plotlist = foim_plots,
  labels = densities,
  ncol = 2
)
```