---
title: "Variation"
output: html_document
vignette: >
  %\VignetteIndexEntry{Variation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(malariasimulation)
library(cowplot)
library(ggplot2)
```

## Variation in outputs

Malariasimulation is a stochastic model, so there will be be variation in model outputs. For example, we could look at detected malaria cases over a year...

```{r}
# A small population
p <- get_parameters(list(
  human_population = 1000,
  individual_mosquitoes = FALSE
))
p <- set_equilibrium(p, 2)
small_o <- run_simulation(365, p)
p <- get_parameters(list(
  human_population = 10000,
  individual_mosquitoes = FALSE
))
p <- set_equilibrium(p, 2)
big_o <- run_simulation(365, p)

plot_grid(
  ggplot(small_o) + geom_line(aes(timestep, n_detect_730_3650)) + ggtitle('n_detect at n = 1,000'),
  ggplot(small_o) + geom_line(aes(timestep, p_detect_730_3650)) + ggtitle('p_detect at n = 1,000'),
  ggplot(big_o) + geom_line(aes(timestep, n_detect_730_3650)) + ggtitle('n_detect at n = 10,000'),
  ggplot(big_o) + geom_line(aes(timestep, p_detect_730_3650)) + ggtitle('p_detect at n = 10,000')
)
```

The n_detect output shows the result of sampling individuals who would be detected by microscopy.

While the p_detect output shows the sum of probabilities of detection in the population.

Notice that the p_detect output is slightly smoother. That's because it forgoes the sampling step. At low population sizes, p_detect will be smoother than the n_detect counterpart.

## Estimating variation 

We can estimate the variation in the number of detectable cases by repeating the simulation several times...

```{r}
outputs <- run_simulation_with_repetitions(
  timesteps = 365,
  repetitions = 10,
  overrides = p,
  parallel=TRUE
)

df <- aggregate(
  outputs$n_detect_730_3650,
  by=list(outputs$timestep),
  FUN = function(x) {
    c(
      median = median(x),
      lowq = unname(quantile(x, probs = .25)),
      highq = unname(quantile(x, probs = .75)),
      mean = mean(x),
      lowci = mean(x) + 1.96*sd(x),
      highci = mean(x) - 1.96*sd(x)
    )
  }
)

df <- data.frame(cbind(t = df$Group.1, df$x))

plot_grid(
  ggplot(df)
  + geom_line(aes(x = t, y = median, group = 1))
  + geom_ribbon(aes(x = t, ymin = lowq, ymax = highq), alpha = .2)
  + xlab('timestep') + ylab('n_detect') + ggtitle('IQR spread'),
  ggplot(df)
  + geom_line(aes(x = t, y = mean, group = 1))
  + geom_ribbon(aes(x = t, ymin = lowci, ymax = highci), alpha = .2)
  + xlab('timestep') + ylab('n_detect') + ggtitle('95% confidence interval')
)

```

These are useful techniques for comparing the expected variance from model outputs to outcomes from trials with lower population sizes.