---
title: "Vector Control: Bednets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector Control: Bednets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(malariasimulation)
library(malariaEquilibrium)
```

Long-lasting insecticidal bednets are highly effective and relatively cheap interventions to prevent malaria. The effects of insecticidal bed net distribution campsigns can be simulated using malariasimulation. The model provides the user with the flexibility to specify parameters that describe the net distribution campaign (e.g. timing, coverage, and target population), the bed nets distributed (e.g. efficacy, longevity), and the relative densities of the mosquito species present in the target population. We will illustrate this through an example with two bednet distributions, once per year.

### Plotting functions
We can create a few plotting functions to visualise the output.
```{r}
cols <- c("#E69F00", "#56B4E9", "#009E73", 
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Plotting functions
plot_prev <- function() {
  plot(x = output$timestep, y = output$n_detect_730_3650/output$n_730_3650, 
       type = 'l', col = cols[3], lwd = 2.5,
       xlab = 'Time (days)', ylab = expression(paste(italic(Pf),"PR"[2-10])),
       xaxs = "i", yaxs = "i", bty = "l", ylim = c(0.15, 0.93))
  lines(x = output_control$timestep, y = output_control$n_detect_730_3650/output_control$n_730_3650,
        col = cols[5], lwd = 2.5)
  abline(v = bednetstimesteps, col = 'black', lty = 2, lwd = 2.5)
  grid(lty = 2, col = 'grey80', lwd = 1)
  legend('bottomleft', box.lty = 0, legend = c('Bednets','Prevalence for bednet scenario','Prevalence for control scenario'),
         col = c('black', cols[3], cols[5]), lty = c(2,1,1), lwd = 2.5)
}
```
  
### Parameterisation

Use the `get_parameters()` function to generate the list of parameters for a seasonal profile, accepting the default values to run the simulation from an equilibrium starting point. The seasonality will drive mosquito dynamics in the absence of interventions. 

```{r}
year <- 365
sim_length <- 3 * year
human_population <- 1000
starting_EIR <- 50

simparams <- get_parameters(
  list(
    human_population = human_population,
    # seasonality parameters
    model_seasonality = TRUE, # Let's try a bi-modal model
    g0 = 0.28605,
    g = c(0.20636, -0.0740318, -0.0009293),
    h = c(0.173743, -0.0730962, -0.116019),
    # Here, we are asking for the output to contain information on severe incidence in children 2-10 years old.
    clinical_incidence_rendering_min_ages = 2*year,
    clinical_incidence_rendering_max_ages = 10*year
  )
)

# Setting 3 species of mosquitoes, 25% An. arabiensis, 25% An. funestus, and 50% An. gambiae
simparams <- set_species(
  simparams, 
  species = list(arab_params, fun_params, gamb_params),
  proportions = c(0.25, 0.25, 0.5)
)

simparams <- set_equilibrium(simparams, starting_EIR)

output_control <- run_simulation(sim_length, simparams)
```

## Simulation

Having established a base set of parameters, we can now create a copy of this parameter list and update it to specify a net distribution campaign. In the example below, we distribute bednets once a year for two years, the first round to 80% of the population and the second round to 50% of the population. It is possible to change the characteristics of the bednets for each distribution timestep by modifying the values for `dn0`, `rn`, `rnm`, and `gamman`. Because there are 3 mosquito species specified in `set_species()` above and two timesteps when nets are distributed, the matrices for `dn0`, `rn`, and `rnm` have 3 columns and 2 rows.
  
```{r, fig.align='center', fig.height=4, fig.width=6}
bednetstimesteps <- c(1, 2) * year # The bednets will be distributed at the end of the first and the second year. 

bednetparams <- set_bednets(
  simparams,
  timesteps = bednetstimesteps,
  coverages = c(.8, .5),  # The first round is distributed to 80% of the population and the second round to 50%.
  retention = 5 * year, # Nets are kept on average 5 years
  dn0 = matrix(c(.533, .45), nrow=2, ncol=3), # Matrix of death probabilities for each mosquito species over time
  rn = matrix(c(.56, .5), nrow=2, ncol=3), # Matrix of repelling probabilities for each mosquito species over time
  rnm = matrix(c(.24, .24), nrow=2, ncol=3), # Matrix of minimum repelling probabilities for each mosquito species over time
  gamman = rep(2.64 * 365, 2) # Vector of bednet half-lives for each distribution timestep
)

output <- run_simulation(sim_length, bednetparams)
```

### Plot prevalence
```{r, fig.align='center', fig.height=5, fig.width=7}
plot_prev()
```
  
## Visualisation 

It is important to understand the difference between the input `coverages` argument of `malariasimulation::set_bednets()` and the resulting population bednet usage over time in the model. In the above example, bednets are distributed to a specified percentage of the population at each distribution round; however, the level of bednet usage may be different. 
  
The average population bednet usage will be influenced by:   

* The size and frequency of distributions specified in `set_bednets()`  
* The assumed net retention half life (`gamman`)
* Correlations in the recipients of nets between rounds  
  
The output from `malariasimulation::run_simulation()` has a variable `n_use_net` that shows the number of people using bednets at any given timestep. We can visualise the proportion of the population using bednets over time to understand how bednet usage changes. 

```{r, fig.align='center', fig.height=5, fig.width=7}
output$prop_use_net <- output$n_use_net / human_population

plot(x = output$timestep, y = output$prop_use_net, type = 'l', 
     col = cols[3], lwd = 2.5, ylim = c(0,1),
     xlab = 'Timestep (days)', ylab = 'Proportion of population using bednets',
     xaxs = "i", yaxs = "i", bty = "l")
grid(lty = 2, col = 'grey80', lwd = 1)
axis(side = 1, lty = 1, col = "black", pos = 0); axis(side = 2, lty = 1, col = "black")
```
  
## Using the `netz` package to estimate coverage inputs needed to achieve target population usage

The [`netz` package](https://mrc-ide.github.io/netz/index.html) is a useful tool to help set up bednets in `malariasimulation`. `malariasimulation` takes as input the proportion of the population who were distributed a bednet at specific time points, but available data for a specific region or country do not always have this information. `netz` allows for simple conversions between bednet metrics like crop, access, distribution, and usage.  
  
See the vignettes from the `netz` package that explain these concepts more comprehensively:  
  
* [This vignette](https://mrc-ide.github.io/netz/articles/overview.html) explains how to estimate the distribution coverage from a specified bednet usage.   
* [This vignette](https://mrc-ide.github.io/netz/articles/population_usage.html) demonstrates how to estimate population bednet usage from an input distribution. 

