---
title: "Vector Control: Spatial Emanator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector Control: Spatial Emanator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
devtools::load_all()
devtools::document()

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=300,
  fig.width=7
)
```

```{r setup}
# Load the requisite packages:
library(malariasimulation)
library(malariaEquilibrium)
# Set colour palette:
cols <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

Spatial emanators can be used both indoors or outdoors. These interventions provide protection through devices that release volatile insecticides that apparently confuse adult female Anopheles mosquitoes. Exposure is understood to "disarm" vectors as they seek blood meals in the proximity. 

`malariasimulation` can be used to investigate the effect of malaria control strategies that deploy spatial emanators indoors or outdoors or in both spaces. Spatial emanators are thought to have mortality inducing effects (Knock down is observed in laboratory tests), repellence effects (measured in laboratory, experimental hut trials and semi-field trials) and effects that seem to confuse the mosquito rendering her less likely to bite should she land. 

The `set_spatial_emanator()` function can be used to parameterise the efficacy, time and coverage of indoor campaigns. 

The corresponding `set_spatial_emanator_outdoor()` function enables outdoor mortality and repellence impacts. 

Finally, the `set_semiochemical` function can be used to introduce a disarming effect to be set that delays the completion of a feed during a gonotrophic cycle.

We will create a few plotting functions to visualise the output.
```{r}
# Plotting functions
plot_prev <- function() {
  plot(x = output_indoors$timestep, y = output_indoors$n_detect_lm_730_3650 / output_indoors$n_age_730_3650, 
       type = "l", col = cols[3], lwd = 1,
       xlab = "Time (days)", ylab = expression(paste(italic(Pf),"PR"[2-10])),
       xaxs = "i", yaxs = "i", ylim = c(0,1))
  
  lines(x = output_outdoors$timestep, y = output_outdoors$n_detect_lm_730_3650 / output_outdoors$n_age_730_3650,
        col = cols[2], lwd = 1)
  lines(x = output_indoors_outdoors$timestep, y = output_indoors_outdoors$n_detect_lm_730_3650 / output_indoors_outdoors$n_age_730_3650,
        col = cols[7], lwd = 1)
  lines(x = output_control$timestep, y = output_control$n_detect_lm_730_3650 / output_control$n_age_730_3650,
        col = cols[5], lwd = 1)
  abline(v = spatial_emanator_timesteps, lty = 2, lwd = 1, col = "black")
  text(x = spatial_emanator_timesteps + 10, y = 0.05, 
       labels = "Emanator\nint.", adj = 0, cex = 0.8)
  grid(lty = 2, col = "grey80", lwd = 0.5)
  legend("topleft", box.lty = 0, 
         legend = c("Prevalence for indoor SE scenario",
                    "Prevalence for outdoor SE scenario",
                    "Prevalence for in & outdoor SE scenario",
                    "Prevalence for control scenario"),
         col = c(cols[3], cols[2],cols[7], cols[5]), lty = c(1,1,1,1), lwd = 2, cex = 0.8, y.intersp = 1.3)
}
```

## Setting Spatial Emanator parameters

### Parameterisation 

Use the `get_parameters()` function to generate a list of parameters, accepting most of the default values, but modifying seasonality values to model a seasonal setting. Then, we use the `set_equilibrium()` function to to initialise the model at a given entomological innoculation rate (EIR). 

```{r}
year <- 365
month <- 30
sim_length <- 3 * year
human_population <- 10000
starting_EIR <- 10

simparams <- get_parameters(
  list(
    human_population = human_population,
    # seasonality parameters
    model_seasonality = TRUE, 
    g0 = 0.285277,
    g = c(-0.0248801, -0.0529426, -0.0168910),
    h = c(-0.0216681, -0.0242904, -0.0073646)
  )
)

simparams <- set_equilibrium(parameters = simparams, init_EIR = starting_EIR)

# Running simulation with no Spatial Emanators
output_control <- run_simulation(timesteps = sim_length, parameters = simparams)
```

#### A note on mosquito species  
It is also possible to use the `set_species()` function to account for 3 different mosquito species in the simulation. In this case, the matrices would need to have additional column corresponding to each mosquito species. For example, if we specified that there were 3 species of mosquitoes in the model and nets were distributed at two timesteps, then the matrices would have 2 rows and 3 columns. If you are not already familiar with the `set_species()` function, see the [Mosquito Species](https://mrc-ide.github.io/malariasimulation/articles/SetSpecies.html) vignette.  

The default parameters are set to model *Anopheles gambiae*. 
```{r}
simparams$species
simparams$species_proportions
```


### Simulation 

Then we can run the simulations for a variety of spatial emanator strategies. In the example below, there are two rounds of spatial emanator, the first at 30% coverage and the second at 80% coverage, each implemented at day 1 of the implementation year. The matrices specified for the parameters for `rse_out_theta`, `rse_out_gamma`, `rse_in_theta`, and `rse_in_gamma` have two rows, one for each timestep where spatial emanator is implemented, and a number of columns corresponding to mosquito species. In this example, we only have 1 column because the species is set to "gamb" as we saw above.

The structure for the spatial emanator model will be published with Kuipou et al. currently in preparation.

#### Example to simulate indoor spatial emanators
```{r, fig.align = 'center', out.width='100%'}

spatial_emanator_timesteps = c(1) * year + 1 # Each round of spatial emanator is implemented in the 1st and second year.

spatial_emanator_params <- set_spatial_emanator(
  simparams,
  timesteps = spatial_emanator_timesteps,
  coverages = c(0.8), # # The first round covers 30% of the population and the second covers 80%. 
  rse_in_theta = matrix(2.025, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of repellence impact parameters indoors
  rse_in_gamma = matrix(-0.009, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of repellence impact parameters indoors
    dse_in_theta = matrix(1.1, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of mortality impact parameters indoors
  dse_in_gamma = matrix(-0.009, nrow=length(spatial_emanator_timesteps), ncol=1) # Matrix of mortality impact parameters indoors

)

# Spatial emanators may also cause a longer gap between feeding attempts by disarming female vectors. The transmission model includes these by extending the foraging time
sim_disarm_params = set_semiochemical(spatial_emanator_params,
                              semiochemical_effect = matrix(c(rep(4.2,length(spatial_emanator_timesteps)),1),
                                                            nrow = length(spatial_emanator_timesteps)+1,
                                                            ncol = 1),
                              timesteps = c(spatial_emanator_timesteps,
800))

output_indoors <- run_simulation(timesteps = sim_length, parameters = sim_disarm_params)
```

#### Example to simulate outdoor spatial emanators
```{r, fig.align = 'center', out.width='100%'}

spatial_emanator_outdoors_params <- set_spatial_emanator_outdoor(
  simparams,
  timesteps = spatial_emanator_timesteps,
  coverages = c(0.8), # # The first round covers 30% of the population and the second covers 80%. 
  rse_out_theta = matrix(1.2, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of impact parameters outdoors
  rse_out_gamma = matrix(-0.009, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of repellence impact parameters outdoors
    dse_out_theta = matrix(1.1, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of mortality impact parameters outdoors
  dse_out_gamma = matrix(-0.009, nrow=length(spatial_emanator_timesteps), ncol=1) # Matrix of mortality impact parameters outdoors
  
)

# Spatial emanators may also cause a longer gap between feeding attempts by disarming female vectors. The transmission model includes these by extending the foraging time
sim_out_disarm_params = set_semiochemical(spatial_emanator_outdoors_params,
                              semiochemical_effect = matrix(c(rep(4.2,length(spatial_emanator_timesteps)),1),
                                                            nrow = length(spatial_emanator_timesteps)+1,
                                                            ncol = 1),
                              timesteps = c(spatial_emanator_timesteps,
800))

output_outdoors <- run_simulation(timesteps = sim_length, parameters = sim_out_disarm_params)
```

#### Example to simulate in and outdoor spatial emanators
```{r, fig.align = 'center', out.width='100%'}

spatial_emanator_everywhere_params <- set_spatial_emanator_outdoor(
  sim_disarm_params,
  timesteps = spatial_emanator_timesteps,
  coverages = c(0.8), # # The first round covers 30% of the population and the second covers 80%. 
  rse_out_theta = matrix(1.2, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of impact parameters outdoors
  rse_out_gamma = matrix(-0.009, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of repellence impact parameters outdoors
    dse_out_theta = matrix(1.1, nrow=length(spatial_emanator_timesteps), ncol=1), # Matrix of mortality impact parameters outdoors
  dse_out_gamma = matrix(-0.009, nrow=length(spatial_emanator_timesteps), ncol=1) # Matrix of mortality impact parameters outdoors
  
)

output_indoors_outdoors <- run_simulation(timesteps = sim_length, parameters = spatial_emanator_everywhere_params)
```
#### Visualisation

```{r, fig.align = 'center', out.width='100%'}
# Plot prevalence
plot_prev()
```

