---
title: "Vector Control: Indoor Residual Spraying"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vector Control: IRS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(malariasimulation)
library(malariaEquilibrium)
```

Indoor Residual Spraying (IRS) involves periodically treating indoor walls with insecticides to eliminate adult female mosquitoes that rest indoors. `malariasimulation` can be used to investigate the effect of malaria control strategies that deploy IRS. Users can set IRS in the model using the `set_spraying()` function to parameterise the time and coverage of spraying campaigns. 

### Plotting functions
We will create a few plotting functions to visualise the output.
```{r}
cols <- c("#E69F00", "#56B4E9", "#009E73", 
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Plotting functions
plot_prev <- function() {
  plot(x = output$timestep, y = output$n_detect_730_3650 / output$n_730_3650, 
       type = "l", col = cols[3], lwd = 2.5,
       xlab = "Time (days)", ylab = expression(paste(italic(Pf),"PR"[2-10])),
       xaxs = "i", yaxs = "i", bty = "l", ylim = c(0,1))
  lines(x = output_control$timestep, y = output_control$n_detect_730_3650 / output_control$n_730_3650,
        col = cols[5], lwd = 2.5)
  abline(v = sprayingtimesteps, lty = 2, lwd = 2.5, col = "black")
  grid(lty = 2, col = "grey80", lwd = 1)
  legend("bottomleft", box.lty = 0, legend = c("Spraying", "Prevalence for IRS scenario","Prevalence for control scenario"),
         col = c("black", cols[3], cols[5]), lty = c(2,1,1), lwd = 2.5)
}
```

### Parameterisation 

Use the `get_parameters()` function to generate a list of parameters, accepting most of the default values, but modifying seasonality values to model a seasonal setting. Then, we use the `set_equilibrium()` function to to initialise the model at a given entomological innoculation rate (EIR). 

```{r}
year <- 365
month <- 30
sim_length <- 3 * year
human_population <- 1000
starting_EIR <- 50

simparams <- get_parameters(
  list(
    human_population = human_population,
    # seasonality parameters
    model_seasonality = TRUE, 
    g0 = 0.285277,
    g = c(-0.0248801, -0.0529426, -0.0168910),
    h = c(-0.0216681, -0.0242904, -0.0073646)
  )
)

simparams <- set_equilibrium(parameters = simparams, init_EIR = starting_EIR)

# Running simulation with no IRS
output_control <- run_simulation(timesteps = sim_length, parameters = simparams)
```

#### A note on mosquito species
It is also possible to use the `set_species()` function to account for 3 different mosquito species in the simulation. In this case, the matrices would need to have additional column corresponding to each mosquito species. For example, if we specified that there were 3 species of mosquitoes in the model and IRS was implemented at two timesteps, then the matrices would have 2 rows and 3 columns. If you are not already familiar with the `set_species()` function, see the Set Species vignette.  

The default setting is to model all mosquito species together. 
```{r}
simparams$species
```


## Simulation 

Then we can run the simulations for a variety of IRS strategies. In the example below, there are two rounds of IRS, the first at 30% coverage and the second at 80% coverage, each 3 months prior to peak rainfall for that year. The function `peak_season_offset()` outputs a timestep when peak rainfall will be observed based on the seasonality profile we set above. Notice that the matrices specified for the parameters for `ls_theta`, `ls_gamma`, `ks_theta`, `ks_gamma`, `ms_theta`, and `ms_gamma` have two rows, one for each timestep where IRS is implemented, and a number of columns corresponding to mosquito species. In this example, we only have 1 column because the species is set to "All" as we saw above.

The structure for IRS model is documented in the supplementary information from Table 3 in the Supplementary Information of [Sherrard-Smith et al., 2018](https://doi.org/10.1038/s41467-018-07357-w). Table S2.1 in the Supplementary Appendix of [Sherrard-Smith et al., 2022](https://doi.org/10.1016/S2542-5196(21)00296-5) has parameter estimates for insecticide resistance for IRS.

```{r, fig.align="center", fig.height=4, fig.width=6}
peak <- peak_season_offset(simparams)

sprayingtimesteps <- c(1, 2) * year + peak - 3 * month # A round of IRS is implemented in the 1st and second year 3 months prior to peak transmission.

sprayingparams <- set_spraying(
  simparams,
  timesteps = sprayingtimesteps,
  coverages = c(0.3, 0.8), # # The first round covers 30% of the population and the second covers 80%. 
  ls_theta = matrix(2.025, nrow=length(sprayingtimesteps), ncol=1), # Matrix of mortality parameters; nrows=length(timesteps), ncols=length(species) 
  ls_gamma = matrix(-0.009, nrow=length(sprayingtimesteps), ncol=1), # Matrix of mortality parameters per round of IRS and per species
  ks_theta = matrix(-2.222, nrow=length(sprayingtimesteps), ncol=1), # Matrix of feeding success parameters per round of IRS and per species
  ks_gamma = matrix(0.008, nrow=length(sprayingtimesteps), ncol=1), # Matrix of feeding success parameters per round of IRS and per species
  ms_theta = matrix(-1.232, nrow=length(sprayingtimesteps), ncol=1), # Matrix of deterrence parameters per round of IRS and per species
  ms_gamma = matrix(-0.009, nrow=length(sprayingtimesteps), ncol=1) # Matrix of deterrence parameters per round of IRS and per species
)

output <- run_simulation(timesteps = sim_length, parameters = sprayingparams)
```


### Plot prevalence
```{r,fig.align="center", fig.height=5, fig.width=7}
plot_prev()
```

